### 2021-10-28 ms/ms.Rmd
Start on MS draft

---
title: "canaper: Categorical analysis of neo- and paleo-endemism in R"
author: "Joel Nitta"
date: "10/27/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Joel H. Nitta, Shawn W. Laffan, Brent D. Mishler, and Wataru Iwasaki

# Abstract

Biodiversity has typically been quantified using species richness, but this ignores evolutionary history.
Due to the increasing availability of robust phylogenies, recently methods have been developed that incorporate phylogenetic relationships into quantification of biodiversity.
CANAPE (categorical analysis of neo- and paleo-endemism) is one such method that can provide insight into the evolutionary processes generating biodiversity.
The only currently available software implementing CANAPE is Biodiverse, which is written in perl and can be used either through a graphical user interface (GUI) or user-developed scripts.
However, many researchers, particularly in the fields of ecology and evolutionary biology, use the R programming language to conduct their analyses.
Here, we provide "canaper", a new R package that provides functions to conduct CANAPE in R.
This will allow researchers to conduct all analyses from data import and cleaning through CANAPE within R, thereby averting the need to manually import and export data and analysis results between programs.
Furthermore, canaper implements methods for efficient computation, including parallelization and encoding of community data as sparse matrices.

# Introduction

Quantifying biodiversity is a major goal of ecology.
Until recently, quantification of biology was done on a purely taxonomic basis: counting the number of species in an area, for example.
However, as all taxa are related to some degree by descent from a common ancestor, a thorough understanding of biodiversity is only possible by taking into account their evolutionary relationships.
This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity (PD) and phylogenetic endemism (PE).
Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies.

One recently developed extension of PE is categorical analysis of neo-and paleo-endemism (CANAPE).
CANAPE uses phylogenetic methods to infer the evolutionary processes that give rise to certain patterns of endemism.
In theory, endemic areas (areas with high rates of range-limited taxa; or in the phylogenetic sense, range-limited portions of the phylogenetic tree) may arise via two major processes: previously wide-spread lineages may undergo extinction in all but a portion of their range (paleo-endemism), or recently diverged lineages may only occur in a small area (neo-endemism).
It is also possible that a given area is home to both paleo- and neo-endemic lineages (mixed endemism).
CANAPE involves analyzing observed patterns of phylogentic endemism in comparison with a null model to infer whether areas are paleo-endemic, neo-endemic, mixed, or lack significantly high endemism.
CANAPE has been used in \> publications, and is a central component of the field of spatial phylogenetics.

Despite the popularity of CANAPE, it has so far only been implemented in one piece of software, Biodiverse.
Biodiverse is written in perl and includes a graphical user interface (GUI).
While Biodiverse is convenient for non-coders because of its GUI, many ecologists and evolutionary biologists use R for their analyses.
R is particularly well-suited for reproducible workflows and is extremely versatile thanks to the thousands of community-contributed packages that extend its utility.
Until now, an R user who wanted to conduct CANAPE analysis as part of a broader workflow needed to first clean raw data in R, export it to Biodiverse, conduct PD and PE analyses in Biodiverse, then import the results back into R for further analysis and visualization.
Besides being error-prone due to its manual nature, this workflow is also not suited to parallel processing, which is needed for large datasets.
Previously, arallel processing and automation of such analyses was only possible using perl scripts; a set of R scripts is available to call perl from R (<https://github.com/NunzioKnerr/biodiverse_pipeline>), but it is not an R package and is not straightforward to use.

Here, we develop a new R package that implements CANAPE completely in R, canaper.
We strive to make canaper simple to use and efficient.
Parallel computing can be enabled with a single line of code.
Randomization algorithms are provided via the `vegan` package, with comes with \> 30 pre-defined algorithms as well as the option to provide a custom, user-defined algorithm.
canaper enables researchers using R to conduct CANAPE completely within R.
canaper has undergone thorough code review through ROpenSci and is verified against a large number of unit tests (100% coverage).
All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.

# Input data format

## Community data

Community data is typically provided as a data.frame or matrix, with species as columns and sites (communities) as rows.
In this case, the data must include both row names and column names.
Community data may also be input as a tibble, in which case site names must be indicated in a dedicated column (default column name "site"), rather than row names since tibbles lack row names.
Community data may be either presence-absence data (0s or 1s) or abundance data (integers \>= 0).
However, all calculations of PD and PE only use presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.

## Phylogeny

The ape R package is used to handle phylogenies, which are stored as lists of the class "phylo".
Phylogenies can be loaded with the ape::read.tree() function.
Phylogenies should have no negative branch lengths, but are not required to be fully bifurcating.

# Analysis workflow

The entire CANAPE workflow can be run with a single function, `cpr_rand_test()`.
However, internally this entails several steps that the user should be aware of as follows.

## Calculate observed values

First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (PD_obs) and phylogenetic endemism (PE_obs) are calculated.
Next, an alternative phylogeny is constructed that has non-zero branch lengths each set to an equal value, then rescaled to a total length of 1.
PD and PE are then calculated on the alternative phylogeny (PD_alt, PE_alt).
Relative PD and PE, the ratio of PD_obs to PD_alt and PE_obs to PE_alt, are then calculated (RPD, RPE).
In CANAPE, areas with extremely high or low RPE are designated as neo- or paleo-endemic, respectively.
However, the statistical significance of any given RPE value cannot be determined from observed values alone.

## Generate random communities

PD (and by extension, PE, RPD, and RPE) is expected to increase with taxon richness, since there is less and less chance of drawing two distantly related taxa without replacement as richness increases.
To determine the statistical significance of RPE, the observed value is compared to a distribution of values obtained from a set of random communities that each have the same richness as the original data.
The random communities are generated by a randomization algorithm that modifies the original data.

Since the randomization algorithm determines the reference (expected) value, the choice of randomization algorithm is likely to have a large effect on the results.
However, there is no single "correct" algorithm, so we have opted to provide the user with a wide range of options.
canaper uses vegan for community matrix randomizations.
vegan includes \>30 randomization algorithms, but not all are appropriate for CANAPE.
Recommended algorithms include `swap` and `curveball` These algorithms preserve column sums (total abundance of each species) and row sums (richness of each site), and both produce results comparable to the randomization algorithm in Biodiverse, rand_structured (see Example).

We have also provided a method for users to provide a custom, user-defined randomization algorithm using the vegan framework.
This may be appropriate if, for example, the community matrix includes a very wide area and it is desired to restrict randomizations to subsets of the area.

## Calculate summary statistics

Once a randomization algorithm has been selected, random communities are generated for a number of replicates set by the user, and a set of summary statistics are computed.
These include the mean and standard error of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the the random communities including standard effect size and rank, which is used to calculate P-value.

To maximize computing efficiency, the individual random communities are not stored in memory.
Since 100s to 1000s or more random replicates may be required, the memory usage can be quite high for large matrices.
Rather, as each random community is generated, the summary statistics are computed immediately and stored.
This way, for any input matrix of size N x M and nr random replicates, where N is number of rows and M is number columns, only nr summary matrices of size N x 1 are stored, rather than nr matrices of size N x M.
This is a key difference of canaper from biodiverse, and may help with computation of large datasets.

## Categorize endemism

# Major functions

## cpr_rand_comm()

## cpr_rand_test()

## cpr_classify_endem()

# Parallel computing

The future package is used for parallel computing.

# Example: *Acacia* in Australia

# Benchmarks

# Conclusion

### 2022-08-30 ms/ms.Rmd
Update MS

@@
 output: html_document
+bibliography: references.yaml
 editor_options: 
@@
 
-Biodiversity has typically been quantified using species richness, but this ignores evolutionary history.
+1. Biodiversity has typically been quantified using species richness, but this ignores evolutionary history.
 Due to the increasing availability of robust phylogenies, recently methods have been developed that incorporate phylogenetic relationships into quantification of biodiversity.
@@
 However, many researchers, particularly in the fields of ecology and evolutionary biology, use the R programming language to conduct their analyses.
-Here, we provide "canaper", a new R package that provides functions to conduct CANAPE in R.
-This will allow researchers to conduct all analyses from data import and cleaning through CANAPE within R, thereby averting the need to manually import and export data and analysis results between programs.
-Furthermore, canaper implements methods for efficient computation, including parallelization and encoding of community data as sparse matrices.
+
+2. Here, we provide canaper, a new R package that provides functions to conduct CANAPE in R.
+canaper implements methods for efficient computation, including parallelization and encoding of community data as sparse matrices.
+The interface is designed for maximum simplicity and reproducibility; CANAPE can be conducted with two functions, and parallel computing can be enabled with one line of code.
+
+3. Our case study shows that canaper produces equivalent results to Biodiverse.
+
+4.
+canaper allows researchers to conduct all analyses from data import and cleaning through CANAPE within R, thereby averting the need to manually import and export data and analysis results between programs.
+We anticipate canaper will become a part of the toolkit for analyzing biodiversity in R.
+
+Key-words: biodiversity, CANAPE, endemism, phylogeny, R
 
@@
 However, as all taxa are related to some degree by descent from a common ancestor, a thorough understanding of biodiversity is only possible by taking into account their evolutionary relationships.
-This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity (PD) and phylogenetic endemism (PE).
+This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity (PD) [@Faith1992] and phylogenetic endemism (PE).
 Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies.
@@
 Besides being error-prone due to its manual nature, this workflow is also not suited to parallel processing, which is needed for large datasets.
-Previously, arallel processing and automation of such analyses was only possible using perl scripts; a set of R scripts is available to call perl from R (<https://github.com/NunzioKnerr/biodiverse_pipeline>), but it is not an R package and is not straightforward to use.
+Previously, parallel processing and automation of such analyses was only possible using perl scripts; a set of R scripts is available to call perl from R (<https://github.com/NunzioKnerr/biodiverse_pipeline>), but it is not an R package and is not straightforward to use.
 
@@
 Randomization algorithms are provided via the `vegan` package, with comes with \> 30 pre-defined algorithms as well as the option to provide a custom, user-defined algorithm.
-canaper enables researchers using R to conduct CANAPE completely within R.
-canaper has undergone thorough code review through ROpenSci and is verified against a large number of unit tests (100% coverage).
+canaper has undergone thorough code review through ROpenSci and is verified against a large number of unit tests (>98% coverage).
 All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.
@@
 
-Community data is typically provided as a data.frame or matrix, with species as columns and sites (communities) as rows.
+Community data is provided as a data.frame or matrix, with species as columns and sites (communities) as rows.
 In this case, the data must include both row names and column names.
@@
 However, all calculations of PD and PE only use presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
+Community data is typically loaded using read.csv, read_csv, or other such functions that can import rectangular data.
 
@@
 The ape R package is used to handle phylogenies, which are stored as lists of the class "phylo".
+Phylogenies should have no negative branch lengths, but are not required to be fully bifurcating.
 Phylogenies can be loaded with the ape::read.tree() function.
-Phylogenies should have no negative branch lengths, but are not required to be fully bifurcating.
 
@@
 
-The entire CANAPE workflow can be run with a single function, `cpr_rand_test()`.
+The entire CANAPE workflow can be run with two functions, cpr_rand_test() and cpr_classify_endem().
 However, internally this entails several steps that the user should be aware of as follows.
@@
 
-PD (and by extension, PE, RPD, and RPE) is expected to increase with taxon richness, since there is less and less chance of drawing two distantly related taxa without replacement as richness increases.
+PD (and by extension, PE, RPD, and RPE) is expected to increase with taxon richness, since there is increasingly less chance of drawing two distantly related taxa without replacement as richness increases.
 To determine the statistical significance of RPE, the observed value is compared to a distribution of values obtained from a set of random communities that each have the same richness as the original data.
-The random communities are generated by a randomization algorithm that modifies the original data.
+The random communities are generated by a randomization algorithm that shuffles the original data.
 
-Since the randomization algorithm determines the reference (expected) value, the choice of randomization algorithm is likely to have a large effect on the results.
-However, there is no single "correct" algorithm, so we have opted to provide the user with a wide range of options.
-canaper uses vegan for community matrix randomizations.
-vegan includes \>30 randomization algorithms, but not all are appropriate for CANAPE.
-Recommended algorithms include `swap` and `curveball` These algorithms preserve column sums (total abundance of each species) and row sums (richness of each site), and both produce results comparable to the randomization algorithm in Biodiverse, rand_structured (see Example).
+Since the randomization algorithm determines the range of reference (expected) values, the choice of randomization algorithm is likely to have a large effect on the results.
+As there is no single "correct" algorithm, we have opted to provide the user with a wide range of options by implementing randomization algorithms included in the vegan package.
+vegan includes \>30 randomization algorithms, but not all are appropriate for CANAPE (not all maintain richness per site).
+Recommended algorithms include `swap` and `curveball`.
+These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to the randomization algorithm in Biodiverse, rand_structured (see Example).
 
@@
 Once a randomization algorithm has been selected, random communities are generated for a number of replicates set by the user, and a set of summary statistics are computed.
-These include the mean and standard error of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the the random communities including standard effect size and rank, which is used to calculate P-value.
+Summary statistics include the mean and standard error of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the random communities including standard effect size and rank, which is used to calculate P-value.
 
@@
 Rather, as each random community is generated, the summary statistics are computed immediately and stored.
-This way, for any input matrix of size N x M and nr random replicates, where N is number of rows and M is number columns, only nr summary matrices of size N x 1 are stored, rather than nr matrices of size N x M.
-This is a key difference of canaper from biodiverse, and may help with computation of large datasets.
+This way, for any input matrix of size N x M and n random replicates, where N is number of rows (sites) and M is number columns (species), n summary matrices each with N rows and one column per summary statistic are stored, rather than n matrices of size N x M.
 
 ## Categorize endemism
+
+The final step in CANAPE is to categorize endemism as described in @Mishler2014.
+Briefly, this is done by comparing significance values of summary statistics calculated in the previous step.
+In order to be considered significantly endemic, a given grid (matrix) cell must first have significantly high PE_obs or PE_alt or both.
+If this is true, the grid cell is classified into one of three non-overlapping categories: if the grid cell has significantly high or low RPE, it is considered to be a center of paleo-endemism or neo-endemism, respectively; if RPE is not signficant (but PE_obs, PE_alt, or both are), it is considered a center of mixed endemism.
+Centers of mixed endemism are futher divided based on *p*-value; if PE_obs and PE_alt and both significant at the $\alpha = 0.01$ level, the grid cell is considered a center of super-endemism.
+While Biodiverse support such categorization, this can be done with the `cpr_classify_endem()` function in `canaper`.
 
@@
 
+The `cpr_rand_comm()` function generates a single random community. 
+The first argument, `comm` is a community data frame. 
+The second, `null_model` is the name of one of the predefined randomization algorithms available in `vegan`.
+The remainder of the arguments are particular to specific types of randomization algorithm.
+
+`cpr_rand_comm()` is typically not called by the user directly, but is provided to help users select randomization algorithms and settings.
+
+One feature to be aware of is that randomization algorithms in vegan are classified as either binary or quantitative
+Binary algorithms are designed for binary (i.e., presence-absence) data, and quantitative algorithms are designed for quantitative (i.e., abundance) data.
+Either type of algorithm will accept either type of data, but binary algorithms will convert abundance data to binary and return a binary matrix.
+
+As the calculations of PD and PE in canaper do not take into account abundance (i.e., no abundance weighting is used), identical results will be obtained by either using abundance data or converting abundance data to binary before analysis.
+In this sense, the binary randomization algorithms are appropriate for CANAPE.
+
+The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with CANAPER, the test data from Phylocom:
+
+```{r cpr-rand-comm-demo}
+data(phylocom)
+
+# The original data includes abundance
+phylocom$comm
+
+# Binary null model produces binary output
+cpr_rand_comm(phylocom$comm, "swap", n_iterations = 100)
+```
+
 ## cpr_rand_test()
+
+The `cpr_rand_test()` function carries out calculation of observed values, generation of random communities, and calculation of summary statistics as described above in "Analysis workflow".
+The main arguments to this function are the input community and phylogeny, type of null model, and various settings for the null model.
+For a full list of null models to choose from, run `?vegan::commsim()`.
+It should be noted that the type of null model, number of random communities, and number of swapping iterations performed per null community (for swapping algorithms) all may strongly affect results of `cpr_rand_test()`.
+While it is beyond the scope of this paper to provide a full discussion of null models in ecology, we have provided details about how to explore appropriate null model settings in the "How many randomizations?" vignette (<https://joelnitta.github.io/canaper/articles/how-many-rand.html>).
+
+The output is a dataframe with communities as rows and summary statistics in columns.
+A large number of summary statistics, including all of those needed to calculate CANAPE, are produced.
+For a full explanation of all output columns, run `?cpr_rand_test()`.
+
+The following code demonstrates usage of `cpr_rand_test()`, using the same example dataset as above:
+
+```{r cpr-rand-comm-demo}
+rand_test_results <- cpr_rand_test(
+  phylocom$comm, phylocom$phy, null_model = "curveball")
+
+# Look a subset of the output columns
+rand_test_results[, 1:9]
+```
 
@@
 
+The `cpr_classify_endem()` function classifies endemism types for the output of `cpr_rand_test()` as described above in "Analysis workflow".
+The input is a dataframe including the following columns calculated by `cpr_rand_test()`: `pe_obs_p_upper` (Upper *p*-value comparing observed phylogenetic endemism to random values), `pe_alt_obs_p_upper` (Upper *p*-value comparing observed phylogenetic endemism on alternate tree to random values), and `rpe_obs_p_upper` (Upper *p*-value comparing observed relative phylogenetic endemism to random values).
+The output is the same dataframe, with the column `endem_type` appended.
+Values of `endem_type` include `paleo` (paleoendemic), `neo` (neoendemic), `not significant`,`mixed` (mixed endemism), and `super` (super-endemic).
+
+The following code demostrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()`:
+
+```{r cpr-classify-endem-demo}
+canape_results <- cpr_classify_endem(rand_test_results)
+
+# Look at the endem_type column
+canape_results[, "endem_type", drop = FALSE]
+```
+
+Note that for this small example, not all possible types of `endem_type` are produced.
+
 # Parallel computing
 
-The future package is used for parallel computing.
+Parallel computing is enabled with the `future` package, which has been designed to allow maximum flexibility in the parallel backend selected by the user (e.g., multiple cores on one machine, multiple remote machines, etc.).
+Parallelization is applied to the calculation of summary statistics for each random community, as there are potentially many random communities (typically > 100 for a robust analysis, though this depends on the dataset).
+To use parallel computing, no changes are needed for `cpr_rand_test()` etc.
+Rather, `future` is loaded with `library()`, then a parallel back-end is specified with `plan()`.
+The user is advised to consult the `future` website (<https://future.futureverse.org/>) for more information on specifying a parallel backend.
+
+The following code demonstrates parallel computing:
+
+```{r cpr-rand-comm-parallel-demo}
+library(future)
+
+# Specify two cores running in parallel
+plan(multisession, workers = 2)
+
+# Usage of cpr_rand_test exactly the same as before
+rand_test_results <- cpr_rand_test(
+  phylocom$comm, phylocom$phy, null_model = "curveball")
+```
+
+There will not be a noticeable decrease in computing time since this dataset is so small.
+However, parallelization can greatly decrease computing time for large datasets; more details are available in the "Parallel computing" vignette (<https://joelnitta.github.io/canaper/articles/parallel.html>).
 
@@
 
-# Benchmarks
+To demonstrate usage of `canaper`, we reproduced the analysis of @Mishler2014, which conduced CANAPE on Australian *Acacia*.
+The *Acacia* dataset is included with `canaper`.
+
+The following code demonstrates how to conduct CANAPE analysis on Australian *Acacia* using `canaper`.
+This is quite simple, requiring only three commands (of which `plan()` is only used for increasing speed and is optional):
+
+```{r acacia-demo}
+library(future)
+library(canaper)
+
+# Specify two cores running in parallel
+plan(multisession, workers = 2)
+
+# Run randomization test, returning output as tibble (not a dataframe)
+acacia_rand_res <- cpr_rand_test(
+  acacia$comm, acacia$phy,
+  null_model = "curveball",
+  n_reps = 20, n_iterations = 100000,
+  tbl_out = TRUE
+)
+
+# Classify endemism
+acacia_canape <- cpr_classify_endem(acacia_rand_res)
+```
+
+`canaper` does not include any plotting functions to visualize the results.
+Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
+Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper`.
+
+```{r acacia-plot}
+library(patchwork) # for multipart-panels
+library(tidyverse) # includes ggplot2, other packages for data tidying
+
+# Fist do some data wrangling to make the results easier to plot
+# (add lat/long columns)
+acacia_canape <- acacia_canape |>
+  separate(site, c("long", "lat"), sep = ":") |>
+  mutate(across(c(long, lat), parse_number))
+
+a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
+  geom_tile() +
+  # cpr_endem_cols is a color palette in canaper for endemism colors
+  scale_fill_manual(values = cpr_endem_cols) +
+  guides(
+    fill = guide_legend(title.position = "top", label.position = "bottom")
+  ) +
+  theme(legend.position = "bottom", legend.title = element_blank())
+
+b <- ggplot(
+  acacia_canape,
+  aes(x = pe_alt_obs, y = pe_obs, color = endem_type)
+) +
+  geom_abline(slope = 1, color = "darkgrey") +
+  geom_point() +
+  scale_color_manual(values = cpr_endem_cols) +
+  labs(
+    x = "Phylogenetic endemism on comparison tree",
+    y = "Phylogenetic endemism on actual tree"
+  ) +
+  theme_bw() +
+  theme(legend.position = "none")
+
+a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
+```
+
+The above plot can be compared to Figure 3 of @Mishler2014.
+As in @Mishler2014, grid cells with significant endemism are primarily located on the coasts, with mostly non-significant grid-cells in the interior.
 

### 2022-08-31 ms/ms.Qmd
Refactor to use quarto

--- 2022-08-30 ms/ms.Rmd
@@
 title: "canaper: Categorical analysis of neo- and paleo-endemism in R"
-author: "Joel Nitta"
-date: "10/27/2021"
-output: html_document
 bibliography: references.yaml
-editor_options: 
-  markdown: 
-    wrap: sentence
+csl: methods-in-ecology-and-evolution.csl
+format:
+  docx:
+    toc: false
+    number-sections: false
+    highlight-style: github
 ---
 
-```{r setup, include=FALSE}
+```{r}
+#| label: setup
+#| include: false
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-```{r cpr-rand-comm-demo}
+```{r}
+#| label: cpr-rand-comm-demo
+library(canaper)
+
 data(phylocom)
@@
 
-```{r cpr-rand-comm-demo}
+```{r}
+#| label: cpr-rand-test-demo
 rand_test_results <- cpr_rand_test(
@@
 
-```{r cpr-classify-endem-demo}
+```{r}
+#| label: cpr-classify-endem-demo
 canape_results <- cpr_classify_endem(rand_test_results)
@@
 
-```{r cpr-rand-comm-parallel-demo}
+```{r}
+#| label: cpr-rand-comm-parallel-demo
 library(future)
@@
 
-```{r acacia-demo}
+```{r}
+#| label: acacia-demo
 library(future)
@@
 
-```{r acacia-plot}
+```{r}
+#| label: acacia-plot
 library(patchwork) # for multipart-panels
@@
 
+The color palette used above has been provided in the package as `cpr_endem_cols`, and was selected to be distinguishable regardless of color vision type.
+We recommend using this color palette for maximum accessibility.
+
 # Conclusion
+
+The `canaper` package enables CANAPE completely within R for the first time.
+This will greatly simplify workflows and facilitate reproducibility for researchers studying spatial biodiversity with R.
+Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.

### 2022-08-31 R/process_refs.R

@@ -1,0 +1,23 @@
+# Filter a list of references in YAML format to those cited in an RMD file
+library(tidyverse)
+library(rmdref)
+
+# Filter YAML references from Zotero library.
+# main_library.yaml has been exported from Zotero
+# like this: file -> "export library" -> "Better CSL YAML"
+filter_refs_yaml(
+  rmd_file = "ms/ms.Qmd",
+  yaml_in = "_targets/user/data_raw/main_library.yaml",
+  yaml_out = "ms/references.yaml", silent = TRUE)
+
+# Write out 'aux' file with citation keys to make collection in Zotero
+# so reference data are easier to work with.
+# The collection can be made with the betterbibtex for Zotero plugin:
+# https://retorque.re/zotero-better-bibtex/citing/aux-scanner/
+extract_citations("ms/ms.Qmd") %>%
+  # aux file just consists of lines formatted like
+  # \citation{KEY}
+  # where KEY is the citation key
+  mutate(latex = glue::glue("\\citation{[key]}", .open = "[", .close = "]")) %>%
+  pull(latex) %>%
+  write_lines("working/canaper_ms.aux")

### 2022-09-01 ms/ms.Qmd
Update workflow and ms

@@
 csl: methods-in-ecology-and-evolution.csl
+highlight-style: github
 format:
-  docx:
-    toc: false
-    number-sections: false
-    highlight-style: github
+  docx: default
+  pdf:
+    include-in-header:
+      # keep figures from floating away
+      text: |
+        \usepackage{float}
+        \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
 ---
@@
 #| include: false
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
+library(targets)
+library(tidyverse)
+library(patchwork)
+
+number <- function(...) {
+  scales::number(big.mark = ",", ...)
+}
 ```
@@
 CANAPE (categorical analysis of neo- and paleo-endemism) is one such method that can provide insight into the evolutionary processes generating biodiversity.
-The only currently available software implementing CANAPE is Biodiverse, which is written in perl and can be used either through a graphical user interface (GUI) or user-developed scripts.
+
+2. The only currently available software implementing CANAPE is Biodiverse, which is written in perl and can be used either through a graphical user interface (GUI) or user-developed scripts.
 However, many researchers, particularly in the fields of ecology and evolutionary biology, use the R programming language to conduct their analyses.
 
-2. Here, we provide canaper, a new R package that provides functions to conduct CANAPE in R.
-canaper implements methods for efficient computation, including parallelization and encoding of community data as sparse matrices.
+3. Here, we describe `canaper`, a new R package that provides functions to conduct CANAPE in R.
+`canaper` implements methods for efficient computation, including parallelization and encoding of community data as sparse matrices.
 The interface is designed for maximum simplicity and reproducibility; CANAPE can be conducted with two functions, and parallel computing can be enabled with one line of code.
 
-3. Our case study shows that canaper produces equivalent results to Biodiverse.
+4. Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 15 min to reproduce a canonical study).
 
 4.
-canaper allows researchers to conduct all analyses from data import and cleaning through CANAPE within R, thereby averting the need to manually import and export data and analysis results between programs.
-We anticipate canaper will become a part of the toolkit for analyzing biodiversity in R.
+`canaper` allows researchers to conduct all analyses from data import and cleaning through CANAPE within R, thereby averting the need to manually import and export data and analysis results between programs.
+We anticipate `canaper` will become a part of the toolkit for analyzing biodiversity in R.
 
@@
 Quantifying biodiversity is a major goal of ecology.
-Until recently, quantification of biology was done on a purely taxonomic basis: counting the number of species in an area, for example.
+Until recently, quantification of biology was done on a purely taxonomic basis: counting the number of species in an area, for example [e.g., @Diamond1975].
 However, as all taxa are related to some degree by descent from a common ancestor, a thorough understanding of biodiversity is only possible by taking into account their evolutionary relationships.
-This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity (PD) [@Faith1992] and phylogenetic endemism (PE).
-Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies.
+This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity (PD) [@Faith1992] and phylogenetic endemism [PE, @Faith1992; @Rosauer2009].
+Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies [e.g., @Thornhill2017; @Mishler2020; @Nitta2022a].
 
-One recently developed extension of PE is categorical analysis of neo-and paleo-endemism (CANAPE).
+One recently developed extension of PE is categorical analysis of neo-and paleo-endemism [CANAPE, @Mishler2014].
 CANAPE uses phylogenetic methods to infer the evolutionary processes that give rise to certain patterns of endemism.
@@
 CANAPE involves analyzing observed patterns of phylogentic endemism in comparison with a null model to infer whether areas are paleo-endemic, neo-endemic, mixed, or lack significantly high endemism.
-CANAPE has been used in \> publications, and is a central component of the field of spatial phylogenetics.
+CANAPE has been used in \> 50 publications (Google Scholar query for papers that cite @Mishler2014 and mention "CANAPE"), and is a central component of the field of spatial phylogenetics.
 
-Despite the popularity of CANAPE, it has so far only been implemented in one piece of software, Biodiverse.
+Despite the popularity of CANAPE, it has so far only been implemented in one piece of software, Biodiverse [@Laffan2010].
 Biodiverse is written in perl and includes a graphical user interface (GUI).
-While Biodiverse is convenient for non-coders because of its GUI, many ecologists and evolutionary biologists use R for their analyses.
+While Biodiverse is convenient for non-coders because of its GUI, many ecologists and evolutionary biologists use R for their analyses [@Lai2019].
 R is particularly well-suited for reproducible workflows and is extremely versatile thanks to the thousands of community-contributed packages that extend its utility.
@@
 
-Here, we develop a new R package that implements CANAPE completely in R, canaper.
-We strive to make canaper simple to use and efficient.
+Here, we develop a new R package that implements CANAPE completely in R, `canaper`.
+We strive to make `canaper` simple to use and efficient.
 Parallel computing can be enabled with a single line of code.
-Randomization algorithms are provided via the `vegan` package, with comes with \> 30 pre-defined algorithms as well as the option to provide a custom, user-defined algorithm.
-canaper has undergone thorough code review through ROpenSci and is verified against a large number of unit tests (>98% coverage).
+Randomization algorithms are provided via the `vegan` package [@Oksanen2017], with comes with \> 30 pre-defined algorithms as well as the option to provide a custom, user-defined algorithm.
+`canaper` has undergone thorough code review through ROpenSci [@Boettiger2015] and is verified against a large number of unit tests (>98% coverage).
 All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.
@@
 
-Community data is provided as a data.frame or matrix, with species as columns and sites (communities) as rows.
+Community data is provided as a `data.frame` or matrix, with species as columns and sites (communities) as rows.
 In this case, the data must include both row names and column names.
-Community data may also be input as a tibble, in which case site names must be indicated in a dedicated column (default column name "site"), rather than row names since tibbles lack row names.
+Community data may also be input as a tibble, in which case site names must be indicated in a dedicated column (default column name `"site"`), rather than row names since tibbles lack row names.
 Community data may be either presence-absence data (0s or 1s) or abundance data (integers \>= 0).
 However, all calculations of PD and PE only use presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
-Community data is typically loaded using read.csv, read_csv, or other such functions that can import rectangular data.
+Community data is typically loaded using `read.csv()`, `readr::read_csv()`, or other functions that can import rectangular data.
 
@@
 
-The ape R package is used to handle phylogenies, which are stored as lists of the class "phylo".
+The `ape` R package is used to handle phylogenies, which are stored as lists of the class `phylo`.
 Phylogenies should have no negative branch lengths, but are not required to be fully bifurcating.
-Phylogenies can be loaded with the ape::read.tree() function.
+Phylogenies can be loaded with the `ape::read.tree()` function.
 
@@
 
-The entire CANAPE workflow can be run with two functions, cpr_rand_test() and cpr_classify_endem().
+The entire CANAPE workflow can be run with two functions, `cpr_rand_test()` and `cpr_classify_endem()`.
 However, internally this entails several steps that the user should be aware of as follows.
@@
 
-First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (PD_obs) and phylogenetic endemism (PE_obs) are calculated.
+First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (`PD_obs`) and phylogenetic endemism (`PE_obs`) are calculated.
 Next, an alternative phylogeny is constructed that has non-zero branch lengths each set to an equal value, then rescaled to a total length of 1.
-PD and PE are then calculated on the alternative phylogeny (PD_alt, PE_alt).
-Relative PD and PE, the ratio of PD_obs to PD_alt and PE_obs to PE_alt, are then calculated (RPD, RPE).
+PD and PE are then calculated on the alternative phylogeny (`PD_alt`, `PE_alt`).
+Relative PD and PE, the ratio of `PD_obs` to `PD_alt` and `PE_obs` to `PE_alt` respectively, are then calculated (RPD, RPE).
 In CANAPE, areas with extremely high or low RPE are designated as neo- or paleo-endemic, respectively.
@@
 
-PD (and by extension, PE, RPD, and RPE) is expected to increase with taxon richness, since there is increasingly less chance of drawing two distantly related taxa without replacement as richness increases.
+PD (and by extension, PE, RPD, and RPE) is expected to increase with taxon richness, since the chance of drawing two distantly related taxa without replacement decreases as richness increases.
 To determine the statistical significance of RPE, the observed value is compared to a distribution of values obtained from a set of random communities that each have the same richness as the original data.
@@
 Since the randomization algorithm determines the range of reference (expected) values, the choice of randomization algorithm is likely to have a large effect on the results.
-As there is no single "correct" algorithm, we have opted to provide the user with a wide range of options by implementing randomization algorithms included in the vegan package.
-vegan includes \>30 randomization algorithms, but not all are appropriate for CANAPE (not all maintain richness per site).
+As there is no single "correct" algorithm, we have opted to provide the user with a wide range of options by implementing randomization algorithms included in the `vegan` package.
+`vegan` includes \>30 randomization algorithms, but not all are appropriate for CANAPE.
 Recommended algorithms include `swap` and `curveball`.
-These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to the randomization algorithm in Biodiverse, rand_structured (see Example).
+These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to the randomization algorithm in Biodiverse, `rand_structured` (see "Example: *Acacia* in Australia").
 
-We have also provided a method for users to provide a custom, user-defined randomization algorithm using the vegan framework.
+We have also provided a method for users to provide a custom, user-defined randomization algorithm using the `vegan` framework.
 This may be appropriate if, for example, the community matrix includes a very wide area and it is desired to restrict randomizations to subsets of the area.
@@
 Once a randomization algorithm has been selected, random communities are generated for a number of replicates set by the user, and a set of summary statistics are computed.
-Summary statistics include the mean and standard error of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the random communities including standard effect size and rank, which is used to calculate P-value.
+Summary statistics include the mean and standard error of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the random communities including standard effect size and rank, which is used to calculate *p*-value.
 
@@
 Briefly, this is done by comparing significance values of summary statistics calculated in the previous step.
-In order to be considered significantly endemic, a given grid (matrix) cell must first have significantly high PE_obs or PE_alt or both.
-If this is true, the grid cell is classified into one of three non-overlapping categories: if the grid cell has significantly high or low RPE, it is considered to be a center of paleo-endemism or neo-endemism, respectively; if RPE is not signficant (but PE_obs, PE_alt, or both are), it is considered a center of mixed endemism.
-Centers of mixed endemism are futher divided based on *p*-value; if PE_obs and PE_alt and both significant at the $\alpha = 0.01$ level, the grid cell is considered a center of super-endemism.
-While Biodiverse support such categorization, this can be done with the `cpr_classify_endem()` function in `canaper`.
+In order to be considered significantly endemic, a given grid (matrix) cell must first have significantly high `PE_obs` or `PE_alt` or both.
+If this is true, the grid cell is classified into one of three non-overlapping categories: if the grid cell has significantly high or low RPE, it is considered to be a center of paleo-endemism or neo-endemism, respectively; if RPE is not signficant (but `PE_obs`, `PE_alt`, or both are), it is considered a center of mixed endemism.
+Centers of mixed endemism are futher divided based on *p*-value; if `PE_obs` and `PE_alt` and both significant at the $\alpha = 0.01$ level, the grid cell is considered a center of super-endemism.
 
@@
 The remainder of the arguments are particular to specific types of randomization algorithm.
-
 `cpr_rand_comm()` is typically not called by the user directly, but is provided to help users select randomization algorithms and settings.
@@
 
-As the calculations of PD and PE in canaper do not take into account abundance (i.e., no abundance weighting is used), identical results will be obtained by either using abundance data or converting abundance data to binary before analysis.
+As the calculations of PD and PE in `canaper` do not take into account abundance (i.e., no abundance weighting is used), identical results will be obtained by either using abundance data or converting abundance data to binary before analysis.
 In this sense, the binary randomization algorithms are appropriate for CANAPE.
@@
 For a full list of null models to choose from, run `?vegan::commsim()`.
-It should be noted that the type of null model, number of random communities, and number of swapping iterations performed per null community (for swapping algorithms) all may strongly affect results of `cpr_rand_test()`.
-While it is beyond the scope of this paper to provide a full discussion of null models in ecology, we have provided details about how to explore appropriate null model settings in the "How many randomizations?" vignette (<https://joelnitta.github.io/canaper/articles/how-many-rand.html>).
+It should be noted that the type of null model, number of random communities, and number of swapping iterations performed per random community (for swapping algorithms) all may strongly affect results of `cpr_rand_test()` [or any metric that is based on comparison to a set of random communties; @Gotelli2001a]
+While it is beyond the scope of this paper to provide a full discussion of null models in ecology, we have provided details about how to explore appropriate null model settings with `canaper` in the "How many randomizations?" vignette (<https://joelnitta.github.io/canaper/articles/how-many-rand.html>).
 
@@
 A large number of summary statistics, including all of those needed to calculate CANAPE, are produced.
-For a full explanation of all output columns, run `?cpr_rand_test()`.
+For a full explanation of all output columns, see Box 1 or run `?cpr_rand_test()`.
 
@@
 The output is the same dataframe, with the column `endem_type` appended.
-Values of `endem_type` include `paleo` (paleoendemic), `neo` (neoendemic), `not significant`,`mixed` (mixed endemism), and `super` (super-endemic).
+Values of `endem_type` include `paleo` (paleoendemic), `neo` (neoendemic), `not significant`, `mixed` (mixed endemism), and `super` (super-endemic).
 
-The following code demostrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()`:
+The following code demonstrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()`:
 
@@
 Parallel computing is enabled with the `future` package, which has been designed to allow maximum flexibility in the parallel backend selected by the user (e.g., multiple cores on one machine, multiple remote machines, etc.).
-Parallelization is applied to the calculation of summary statistics for each random community, as there are potentially many random communities (typically > 100 for a robust analysis, though this depends on the dataset).
+Parallelization is applied to the calculation of summary statistics for each random community, as there are potentially many random communities (typically \> 100 for a robust analysis, though this depends on the dataset).
 To use parallel computing, no changes are needed for `cpr_rand_test()` etc.
@@
 #| label: cpr-rand-comm-parallel-demo
+#| eval: false
 library(future)
@@
 
-To demonstrate usage of `canaper`, we reproduced the analysis of @Mishler2014, which conduced CANAPE on Australian *Acacia*.
+To demonstrate usage of `canaper` with a more realistic dataset, we reproduced the analysis of @Mishler2014, which conduced CANAPE on Australian *Acacia*.
 The *Acacia* dataset is included with `canaper`.
+It includes a community matrix comprising `r number(dim(acacia$comm)[2])` species × `r number(dim(acacia$comm)[1])` sites and a phylogeny with `r number(ape::Ntip(acacia$phy))` tips (same species as in the community data plus two outgroup taxa).
 
 The following code demonstrates how to conduct CANAPE analysis on Australian *Acacia* using `canaper`.
-This is quite simple, requiring only three commands (of which `plan()` is only used for increasing speed and is optional):
+This is quite simple, requiring only two commands.
+
+```{r}
+#| label: acacia-load
+#| include: false
+# Load pre-calculated results from targets pipeline
+# - acacia_canape_cpr is Acacia CANAPE results calculated with canaper
+# - rename it so we can use it in demo code below
+acacia_canape <- tar_read(acacia_canape_cpr,
+  store = here::here("_targets"))
+```
 
@@
 #| label: acacia-demo
-library(future)
+#| eval: false
 library(canaper)
 
-# Specify two cores running in parallel
-plan(multisession, workers = 2)
-
-# Run randomization test, returning output as tibble (not a dataframe)
+# Run randomization test, returning output as tibble
 acacia_rand_res <- cpr_rand_test(
@@
   null_model = "curveball",
-  n_reps = 20, n_iterations = 100000,
+  n_reps = 999, n_iterations = 100000,
   tbl_out = TRUE
@@
 Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
-Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper`.
+Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr).
 
 ```{r}
-#| label: acacia-plot
+#| label: acacia-plot-demo
+#| fig.height: 8
+#| eval: false
 library(patchwork) # for multipart-panels
@@
   ) +
+  coord_fixed() +
   theme(legend.position = "bottom", legend.title = element_blank())
@@
 
-The above plot can be compared to Figure 3 of @Mishler2014.
-As in @Mishler2014, grid cells with significant endemism are primarily located on the coasts, with mostly non-significant grid-cells in the interior.
+As in Figure 3 of @Mishler2014, grid cells with significant endemism are primarily located on the coasts, with mostly non-significant grid-cells in the interior (@fig-acacia-cpr).
+Furthermore, the endemism types largely correspond between the two figures.
 
-The color palette used above has been provided in the package as `cpr_endem_cols`, and was selected to be distinguishable regardless of color vision type.
+```{r}
+#| label: compare-cpr-biod
+#| include: false
+
+# Load CANAPE results from canaper and biodiverse
+tar_load(c(
+  acacia_canape_cpr,
+  acacia_canape_biod,
+  acacia_canape_comp),
+  store = here::here("_targets")
+)
+
+# Extract percent agreement
+p_agree <- scales::percent(acacia_canape_comp$p_agree, accuracy = 0.1)
+
+# Make table comparing the two
+comp_table <- left_join(
+  count(acacia_canape_cpr, endem_type, name = "n_canaper"),
+  count(acacia_canape_biod, endem_type, name = "n_biodiverse"),
+  by = "endem_type")
+```
+
+The color palette used in @fig-acacia-cpr has been provided in `canaper` as `cpr_endem_cols`, and was selected to be distinguishable regardless of color vision type [@Okabe2002].
 We recommend using this color palette for maximum accessibility.
+
+We also re-ran the *Acacia* analysis using Biodiverse with the same settings as @Mishler2014 (999 replicates of the `rand_structured` null model) and compared the results with those from `canaper`.
+Importantly, there is no expectation that results between the two should match exactly, for two reasons.
+First, the null model used between the Biodiverse analysis (`rand_structured`) and the `canaper` (`curveball`) differs.
+`rand_structured` is not currently available in R, but we hope to add this to a future version of `canaper`.
+Second, the random communities generated in each run will be different, so the exact *p*-values will also be different.
+With a sufficiently high number of random communities, significance (e.g., at the $\alpha = 0.5$ level) is expected to converge, but there may be borderline cases that appear signfificant in some analyses and non-significant in others.
+
+When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells.
+Differences between the two are mostly due to a slightly greater number of significant sites in the `canaper` results (@tbl-cpr-biod).
+Again, this does not mean that one or the other is more correct; rather, the differences in significant endemism likely reflect differences in the null models that produced the random communities.
+Here, both the `rand_structured` and `curveball` null models preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models, which may explain the relative high agreement between results.
+Selection of an appropriate null model is beyond the scope of this paper, but must be considered carefully in any community ecological analysis.
+
+```{r}
+#| label: build-times
+#| include: false
+
+# Enforce dependency on acacia_canape_cpr_single
+tar_load(acacia_canape_cpr_single,
+  store = here::here("_targets"))
+
+# Load build times as a list
+times <- targets::tar_meta(
+  fields = seconds,
+  store = here::here("_targets")
+  ) %>%
+  mutate(minutes = number(seconds / 60, accuracy = 0.1)) %>%
+  split(.$name)
+```
+
+Computations were carried out on an iMac (2019) with 16 GB RAM and a 3GHz, six core processor.
+Approximate compute times were 1.3 hr for Biodiverse and `r times$acacia_canape_cpr_single$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
+However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr$minutes` min (further increases in number of cores did not significantly decrease time).
+This demonstrates that `canaper` can efficiently compute moderately sized datasets with a personal computer.
+
+# Comparison with other R packages
+
+We are not aware of any other R packages that conduct the entire CANAPE pipeline automatically.
+However, there is a large number of packages for analyzing species diversity [approximately 40 packages out of 15,300 as of 2019\; @Pavoine2020].
+Some of those that are more closely related to `canaper` include the following.
+`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD.
+`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing speed of PD, and is used by `canaper`.
+`vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
+`vegan` includes by far the greatest number of algorithms for generating random communities (> 30), and is used by `canaper`.
+`adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.
+`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics, but does not appear to be currently maintained and has been removed from CRAN.
 
@@
 Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
+We expect `canaper` will become a major tool in the toolkit of the emerging field of spatial phylogenetics alongside Biodiverse.
+
+# References
+<!-- Insert <div id="refs"></div> to place the refs here instead of at the end -->
+<div id="refs"></div>
+
+# Figures
+```{r}
+#| label: fig-acacia-cpr
+#| fig.height: 8
+#| fig-cap: "Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*."
+#| echo: false
+
+# Fist do some data wrangling to make the results easier to plot
+# (add lat/long columns)
+acacia_canape <- acacia_canape |>
+  separate(site, c("long", "lat"), sep = ":") |>
+  mutate(across(c(long, lat), parse_number))
+
+a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
+  geom_tile() +
+  # cpr_endem_cols is a color palette in canaper for endemism colors
+  scale_fill_manual(values = cpr_endem_cols) +
+  guides(
+    fill = guide_legend(title.position = "top", label.position = "bottom")
+  ) +
+  coord_fixed() +
+  theme(legend.position = "bottom", legend.title = element_blank())
+
+b <- ggplot(
+  acacia_canape,
+  aes(x = pe_alt_obs, y = pe_obs, color = endem_type)
+) +
+  geom_abline(slope = 1, color = "darkgrey") +
+  geom_point() +
+  scale_color_manual(values = cpr_endem_cols) +
+  labs(
+    x = "Phylogenetic endemism on comparison tree",
+    y = "Phylogenetic endemism on actual tree"
+  ) +
+  theme_bw() +
+  theme(legend.position = "none")
+
+a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
+```
+
+(a) Map of Australia showing grid cells (communities) colored by endemism type.
+(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.
+a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type.
+This figure reproduces Figure 3 of @Mishler2014.
+
+# Tables
+
+```{r}
+#| label: tbl-cpr-biod
+#| tbl-cap: "Comparison of CANAPE results for Australian *Acacia* between `canaper` and Biodiverse"
+#| echo: false
+
+knitr::kable(comp_table)
+```
+
+# Boxes
+
+**Box 1**: Columns in dataframe produced by `cpr_rand_test()`.
+
+`cpr_rand_test()` calculates four basic metrics: `pd`, phylogenetic diversity (PD) [@Faith1992]; `rpd` [@Mishler2014], relative PD; `pe` [@Rosauer2009], phylogenetic endemism (PE); `rpe` relative PE [@Mishler2014].
+
+Four each of the four basic metrics, 9 columns are calculated:
+
+- `*_obs`: Observed value
+- `*_obs_c_lower`: Count of times observed value was lower than random values
+- `*_obs_c_upper`: Count of times observed value was higher than random values
+- `*_obs_p_lower`: Percentage of times observed value was lower than random values
+- `*_obs_p_upper`: Percentage of times observed value was higher than random values
+- `*_obs_q`: Count of the non-NA random values used for comparison
+- `*_obs_z`: Standard effect size (*z*-score)
+- `*_rand_mean`: Mean of the random values
+- `*_rand_sd`: Standard deviation of the random values
+
+So for `pd` the output columns would include `pd_obs`, `pd_obs_c_lower`, etc.

### 2022-09-01 R/functions.R

@@ -1,0 +1,94 @@
+run_canape <- function(
+  comm, phy, null_model, n_reps = 500, n_iterations = 100000, workers = 6) {
+  on.exit(plan(sequential))
+  # Specify two cores running in parallel
+  plan(multisession, workers = workers)
+  # Run randomization test, returning output as tibble (not a dataframe)
+  rand_res <- cpr_rand_test(
+    comm, phy,
+    null_model = null_model,
+    n_reps = n_reps, n_iterations = n_iterations,
+    tbl_out = TRUE
+  )
+  # Unset parallelization
+  plan(sequential)
+  # Classify endemism
+  res <- cpr_classify_endem(rand_res) %>%
+    mutate(site = str_remove_all(site, "'"))
+  return(res)
+}
+
+#' Test number of optimal iterations needed for a given null model
+test_iter <- function(comm, null_model, n_iter, binary) {
+  if (isTRUE(binary)) {
+    # binary algorithms produce binary matrices, so convert input
+    # matrix to binary too
+    comm <- as.matrix(comm)
+    comm[comm > 0] <- 1
+  }
+  # Use a tibble to hold the data and run loops
+  tibble::tibble(
+    n_iter = n_iter,
+    # Make one random community for each value of `n_iter`
+    rand_comm = purrr::map(
+      n_iter,
+      ~ canaper::cpr_rand_comm(
+        comm,
+        null_model = null_model, n_iterations = .
+      )
+    ),
+    # Calculate the Pearson correlation (r) between each randomized
+    # matrix and the original matrix by converting
+    # the matrices to vectors
+    corr = map_dbl(
+      rand_comm,
+      ~ cor(c(comm), c(.)) # c() converts a matrix to vector
+    )
+  )
+}
+
+#' Classify endemism types for output from Biodiverse
+#'
+#' Description of Biodiverse output columns and corresponding canaper columns
+#' as follows
+#'
+#' PE_WE_P: Phylogenetic weighted endemism as a proportion of the total tree
+#' length
+#'
+#' PHYLO_RPE2: Relative Phylogenetic Endemism (RPE). The ratio of the tree's PE
+#' to a null model where PE is calculated using a tree where all branches are of
+#' equal length.
+#' = rpe_obs_p_upper
+#'
[…]

### 2022-09-01 R/packages.R

@@ -1,0 +1,9 @@
+library(canaper)
+library(future)
+library(tidyverse)
+library(assertr)
+library(glue)
+library(conflicted)
+
+conflict_prefer("filter", "dplyr")
+conflict_prefer("lag", "dplyr")

### 2022-09-01 _targets.R

@@ -1,0 +1,65 @@
+library(targets)
+library(tarchetypes)
+
+source("R/packages.R")
+source("R/functions.R")
+
+tar_plan(
+  # Load acacia dataset
+  acacia = canaper::acacia,
+  # Test optimal number of iterations for making random communities
+  acacia_iter_test_res = test_iter(
+    comm = acacia$comm,
+    null_model = "curveball",
+    # test iterations from 10 to 10^7
+    n_iter = magrittr::raise_to_power(10, 1:7),
+    binary = TRUE
+  ),
+  # Run CANAPE using canaper (2 cores)
+  acacia_canape_cpr = run_canape(
+    comm = acacia$comm,
+    phy = acacia$phy,
+    null_model = "curveball",
+    n_reps = 999,
+    n_iterations = 100000,
+    workers = 2
+  ),
+  # Repeat, with only one core for comparison
+  acacia_canape_cpr_3 = run_canape(
+    comm = acacia$comm,
+    phy = acacia$phy,
+    null_model = "curveball",
+    n_reps = 999,
+    n_iterations = 100000,
+    workers = 3
+  ),
+  acacia_canape_cpr_single = run_canape(
+    comm = acacia$comm,
+    phy = acacia$phy,
+    null_model = "curveball",
+    n_reps = 999,
+    n_iterations = 100000,
+    workers = 1
+  ),
+  # Load results from Biodiverse for Acacia
+  tar_file_read(
+    acacia_biod_res,
+    "_targets/user/data_raw/CANAPE_Acacia/spatial_results_prank.csv",
+    read_csv(!!.x)),
+  # Classify endemism types for Biodiverse results
+  acacia_canape_biod = classify_biod(acacia_biod_res),
+  # Compare % agreement between canaper and Biodiverse
+  acacia_canape_comp = calc_agree_endem(
+    df_1 = acacia_canape_cpr,
+    df_2 = acacia_canape_biod
+  ),
+  # Render MS
+  tar_quarto(
+    ms_doc,
+    "ms/ms.Qmd",
+    extra_files = c(
+      "ms/methods-in-ecology-and-evolution.csl",
+      "ms/references.yaml"
+    )
+  )
+)

### 2022-09-06 ms/ms.Qmd
Add Box with definitions

@@
 
-**Box 1**: Columns in dataframe produced by `cpr_rand_test()`.
+**Box 1**: Glossary of terms and metrics involved in categorical analysis of neo- and paleo-endemism (CANAPE)
 
-`cpr_rand_test()` calculates four basic metrics: `pd`, phylogenetic diversity (PD) [@Faith1992]; `rpd` [@Mishler2014], relative PD; `pe` [@Rosauer2009], phylogenetic endemism (PE); `rpe` relative PE [@Mishler2014].
+- Community: A co-occurring set of species.
+- Grid cell: One of a set of spatial units each with equal area that together cover a region of interest; each grid cell comprises a community.
+- Community matrix: A 2-dimensional dataset with communties (grid cells) on one axis and species on the other.
+- Phylogenetic diversity (PD): The total branch length (including the root) of a phylogenetic tree that includes all species occurring in a given area (community or grid cell) [@Faith1992].
+- Phylogenetic endemism (PE): PD inversely weighted by the distribution range (i.e., number of grid cells) of a branch [@Rosauer2009].
+- Relative phylogenetic diversity (RPD): The ratio of PD measured on the original tree vs.
+PD measured on a comparison tree where all branch lengths have been transformed to equal length [@Mishler2014].
+- Relative phylogenetic endemism (RPE): The ratio of PE measured on the original tree vs.
+PE measured on a comparison tree where all branch lengths have been transformed to equal length [@Mishler2014].
+- Neoendemic: An area with a high concentration of short, range-restricted branchlengths; may be due to processes such as recent speciation (radiation)
+- Paleoendemic: An area with a high concentration of long, range-restricted branchlengths; may be due to processes such as extinction or colonization by distantly related species
+- Mixed endemism: An area with a mixture of short and long, range-restricted branchlengths; may be due to multiple processes
+- Super endemism: An area with mixed endemism that is statistically highly significant
+
+**Box 2**: Columns in dataframe produced by `cpr_rand_test()`.
+
+`cpr_rand_test()` calculates four basic metrics: `pd`, phylogenetic diversity (PD); `rpd` [@Mishler2014], relative PD; `pe` [@Rosauer2009], phylogenetic endemism (PE); `rpe` relative PE [@Mishler2014] (see Box 1 for an explanation of each).
 

### 2022-09-06 ms/ms.Qmd
Update MS

--- 2022-09-01 ms/ms.Qmd
@@
 1. Biodiversity has typically been quantified using species richness, but this ignores evolutionary history.
-Due to the increasing availability of robust phylogenies, recently methods have been developed that incorporate phylogenetic relationships into quantification of biodiversity.
+Due to the increasing availability of robust phylogenies, methods have been developed that incorporate phylogenetic relationships into quantification of biodiversity.
 CANAPE (categorical analysis of neo- and paleo-endemism) is one such method that can provide insight into the evolutionary processes generating biodiversity.
-
-2. The only currently available software implementing CANAPE is Biodiverse, which is written in perl and can be used either through a graphical user interface (GUI) or user-developed scripts.
+The only currently available software implementing CANAPE is Biodiverse, which is written in perl and can be used either through a graphical user interface (GUI) or user-developed scripts.
 However, many researchers, particularly in the fields of ecology and evolutionary biology, use the R programming language to conduct their analyses.
 
-3. Here, we describe `canaper`, a new R package that provides functions to conduct CANAPE in R.
+2. Here, we describe `canaper`, a new R package that provides functions to conduct CANAPE in R.
 `canaper` implements methods for efficient computation, including parallelization and encoding of community data as sparse matrices.
@@
 
-4. Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 15 min to reproduce a canonical study).
+3. Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 15 min to reproduce a canonical study).
 
@@
 
-Key-words: biodiversity, CANAPE, endemism, phylogeny, R
+Key-words: biodiversity, CANAPE, canaper, endemism, phylogeny, R, spatial phylogenetics
 
@@
 However, as all taxa are related to some degree by descent from a common ancestor, a thorough understanding of biodiversity is only possible by taking into account their evolutionary relationships.
-This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity (PD) [@Faith1992] and phylogenetic endemism [PE, @Faith1992; @Rosauer2009].
-Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies [e.g., @Thornhill2017; @Mishler2020; @Nitta2022a].
+This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity [PD\; @Faith1992] and phylogenetic endemism [PE\; @Faith1992; @Rosauer2009] (Box 1).
+Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies [e.g., @Mishler2020; @Nitta2022a; @Thornhill2017].
 
-One recently developed extension of PE is categorical analysis of neo-and paleo-endemism [CANAPE, @Mishler2014].
+One recently developed extension of PE is categorical analysis of neo-and paleo-endemism [CANAPE\; @Mishler2014].
 CANAPE uses phylogenetic methods to infer the evolutionary processes that give rise to certain patterns of endemism.
@@
 While Biodiverse is convenient for non-coders because of its GUI, many ecologists and evolutionary biologists use R for their analyses [@Lai2019].
-R is particularly well-suited for reproducible workflows and is extremely versatile thanks to the thousands of community-contributed packages that extend its utility.
+R is well-suited for reproducible workflows and is extremely versatile thanks to the thousands of community-contributed packages that extend its utility.
 Until now, an R user who wanted to conduct CANAPE analysis as part of a broader workflow needed to first clean raw data in R, export it to Biodiverse, conduct PD and PE analyses in Biodiverse, then import the results back into R for further analysis and visualization.
+Also, while Biodiverse conducted the metrics needed to categorize endemism type, it did not actually perform this categorization.
 Besides being error-prone due to its manual nature, this workflow is also not suited to parallel processing, which is needed for large datasets.
@@
 
-Here, we develop a new R package that implements CANAPE completely in R, `canaper`.
-We strive to make `canaper` simple to use and efficient.
+Here, we describe a new R package that implements CANAPE completely in R, `canaper`.
+We strived to make `canaper` simple to use and efficient.
 Parallel computing can be enabled with a single line of code.
-Randomization algorithms are provided via the `vegan` package [@Oksanen2017], with comes with \> 30 pre-defined algorithms as well as the option to provide a custom, user-defined algorithm.
+Randomization algorithms are provided via the `vegan` package [@Oksanen2017], which comes with \> 30 pre-defined algorithms as well as the option to provide a custom, user-defined algorithm.
 `canaper` has undergone thorough code review through ROpenSci [@Boettiger2015] and is verified against a large number of unit tests (>98% coverage).
@@
 
-Community data is provided as a `data.frame` or matrix, with species as columns and sites (communities) as rows.
+Community data is provided as a `data.frame` or matrix, with species as columns and communities (also referred to as "sites" or "grid-cells") as rows.
 In this case, the data must include both row names and column names.
@@
 
-First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (`PD_obs`) and phylogenetic endemism (`PE_obs`) are calculated.
-Next, an alternative phylogeny is constructed that has non-zero branch lengths each set to an equal value, then rescaled to a total length of 1.
-PD and PE are then calculated on the alternative phylogeny (`PD_alt`, `PE_alt`).
-Relative PD and PE, the ratio of `PD_obs` to `PD_alt` and `PE_obs` to `PE_alt` respectively, are then calculated (RPD, RPE).
-In CANAPE, areas with extremely high or low RPE are designated as neo- or paleo-endemic, respectively.
+First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (`pd_obs`) and phylogenetic endemism (`pe_obs`) are calculated.
+Next, an alternative phylogeny (also referred to as the "comparison tree") is constructed that has non-zero branch lengths each set to an equal value, then rescaled to a total length of 1.
+PD and PE are then calculated on the alternative phylogeny (`pd_alt`, `pe_alt`).
+Relative PD and PE, the ratio of `pd_obs` to `pd_alt` and `pe_obs` to `pe_alt` respectively, are then calculated (RPD, RPE).
+In CANAPE, areas (grid-cells) with extremely high or low RPE are designated as paleo- or neo-endemic, respectively.
 However, the statistical significance of any given RPE value cannot be determined from observed values alone.
@@
 `vegan` includes \>30 randomization algorithms, but not all are appropriate for CANAPE.
-Recommended algorithms include `swap` and `curveball`.
-These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to the randomization algorithm in Biodiverse, `rand_structured` (see "Example: *Acacia* in Australia").
+Recommended algorithms include `swap` [@Gotelli2003] and `curveball` [@Strona2014].
+These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to the randomization algorithm in Biodiverse, `rand_structured` (see "Example: Australian *Acacia*").
 
@@
 
-Once a randomization algorithm has been selected, random communities are generated for a number of replicates set by the user, and a set of summary statistics are computed.
-Summary statistics include the mean and standard error of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the random communities including standard effect size and rank, which is used to calculate *p*-value.
+Once a randomization algorithm has been selected, random communities are generated for a number of replicates set by the user, and a set of summary statistics are computed (Box 2).
+Summary statistics include the mean and standard deviation of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the random communities including standard effect size and rank, which is used to calculate *p*-value.
 
@@
 Rather, as each random community is generated, the summary statistics are computed immediately and stored.
-This way, for any input matrix of size N x M and n random replicates, where N is number of rows (sites) and M is number columns (species), n summary matrices each with N rows and one column per summary statistic are stored, rather than n matrices of size N x M.
+This way, for any input matrix of size $N \times M$ and $n$ random replicates, where $N$ is number of rows (sites) and $M$ is number columns (species), $n$ summary matrices each with $N$ rows and one column per summary statistic are stored, rather than $n$ matrices of size $N \times M$.
 
@@
 The final step in CANAPE is to categorize endemism as described in @Mishler2014.
-Briefly, this is done by comparing significance values of summary statistics calculated in the previous step.
-In order to be considered significantly endemic, a given grid (matrix) cell must first have significantly high `PE_obs` or `PE_alt` or both.
-If this is true, the grid cell is classified into one of three non-overlapping categories: if the grid cell has significantly high or low RPE, it is considered to be a center of paleo-endemism or neo-endemism, respectively; if RPE is not signficant (but `PE_obs`, `PE_alt`, or both are), it is considered a center of mixed endemism.
-Centers of mixed endemism are futher divided based on *p*-value; if `PE_obs` and `PE_alt` and both significant at the $\alpha = 0.01$ level, the grid cell is considered a center of super-endemism.
+Briefly, this is done by comparing significance values of summary statistics calculated in the previous step (Box 3).
+In order to be considered significantly endemic, a given grid-cell must first have significantly high `pe_obs` or `pe_alt` or both.
+If this is true, the grid-cell is classified into one of three non-overlapping categories: if the grid-cell has significantly high or low RPE, it is considered to be a center of paleo-endemism or neo-endemism, respectively; if RPE is not signficant (but `pe_obs`, `pe_alt`, or both are), it is considered a center of mixed endemism.
+Centers of mixed endemism are futher divided based on *p*-value; if `pe_obs` and `pe_alt` and both significant at the $\alpha = 0.01$ level, the grid-cell is considered a center of super-endemism.
 
@@
 
-## cpr_rand_comm()
+## `cpr_rand_comm()`
 
@@
 
-One feature to be aware of is that randomization algorithms in vegan are classified as either binary or quantitative
+One feature to be aware of is that randomization algorithms in vegan are classified as either binary or quantitative.
 Binary algorithms are designed for binary (i.e., presence-absence) data, and quantitative algorithms are designed for quantitative (i.e., abundance) data.
@@
 
-The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with CANAPER, the test data from Phylocom:
+The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with CANAPER, the test data from Phylocom [@Webb2008]:
 
@@
 
-## cpr_rand_test()
+## `cpr_rand_test()`
 
@@
 For a full list of null models to choose from, run `?vegan::commsim()`.
-It should be noted that the type of null model, number of random communities, and number of swapping iterations performed per random community (for swapping algorithms) all may strongly affect results of `cpr_rand_test()` [or any metric that is based on comparison to a set of random communties; @Gotelli2001a]
+It should be noted that the type of null model, number of random communities, and number of swapping iterations performed per random community (for swapping algorithms) all may strongly affect results of `cpr_rand_test()` [or any metric that is based on comparison to a set of random communties\; @Gotelli2001a]
 While it is beyond the scope of this paper to provide a full discussion of null models in ecology, we have provided details about how to explore appropriate null model settings with `canaper` in the "How many randomizations?" vignette (<https://joelnitta.github.io/canaper/articles/how-many-rand.html>).
@@
 A large number of summary statistics, including all of those needed to calculate CANAPE, are produced.
-For a full explanation of all output columns, see Box 1 or run `?cpr_rand_test()`.
+For a full explanation of all output columns, see Box 2 or run `?cpr_rand_test()`.
 
@@
 
-## cpr_classify_endem()
+## `cpr_classify_endem()`
 
 The `cpr_classify_endem()` function classifies endemism types for the output of `cpr_rand_test()` as described above in "Analysis workflow".
-The input is a dataframe including the following columns calculated by `cpr_rand_test()`: `pe_obs_p_upper` (Upper *p*-value comparing observed phylogenetic endemism to random values), `pe_alt_obs_p_upper` (Upper *p*-value comparing observed phylogenetic endemism on alternate tree to random values), and `rpe_obs_p_upper` (Upper *p*-value comparing observed relative phylogenetic endemism to random values).
+The input is a dataframe including the following columns calculated by `cpr_rand_test()`: `pe_obs_p_upper` (upper *p*-value comparing observed phylogenetic endemism to random values), `pe_alt_obs_p_upper` (upper *p*-value comparing observed phylogenetic endemism on alternate tree to random values), and `rpe_obs_p_upper` (upper *p*-value comparing observed relative phylogenetic endemism to random values).
 The output is the same dataframe, with the column `endem_type` appended.
@@
 
-The following code demonstrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()`:
+The following code demonstrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()` (note that for this small example, not all possible types of `endem_type` are produced):
 
@@
 
-Note that for this small example, not all possible types of `endem_type` are produced.
+In addition to `cpr_classify_endem()`, `canaper` includes the `cpr_classify_signif()` function for assessing significance level of PD, RPD, PE, and RPE.
+Both `cpr_classify_endem()` and `cpr_classify_signif()` take a dataframe as their first argument and return a dataframe as output, so they are "pipe friendly", i.e., can be chained together using the pipe (`%>%` or `|>`).
 
@@
 
-# Example: *Acacia* in Australia
+# Datasets
 
-To demonstrate usage of `canaper` with a more realistic dataset, we reproduced the analysis of @Mishler2014, which conduced CANAPE on Australian *Acacia*.
-The *Acacia* dataset is included with `canaper`.
-It includes a community matrix comprising `r number(dim(acacia$comm)[2])` species × `r number(dim(acacia$comm)[1])` sites and a phylogeny with `r number(ape::Ntip(acacia$phy))` tips (same species as in the community data plus two outgroup taxa).
+`canaper` comes with three datasets for testing and demonstration (@tbl-datasets).
+Each is a list with two named elements, `comm` (community matrix, dataframe) and `phy` (phylogenetic tree, list of class `"phylo"`).
+`acacia` includes data of Australian *Acacia* analyzed by @Mishler2014.
+`phylocom` and `biod_example` are both fictitious datasets compiled for testing software, and were originally published in Phylocom [@Webb2008] and Biodiverse [@Laffan2010], respectively.
 
-The following code demonstrates how to conduct CANAPE analysis on Australian *Acacia* using `canaper`.
-This is quite simple, requiring only two commands.
+# Example: Australian *Acacia*
+
+## Analysis with `canaper`
+
+To demonstrate usage with a real dataset, we reproduced the analysis of @Mishler2014, who conducted CANAPE on Australian *Acacia* using Biodiverse.
+Conducting CANAPE on Australian *Acacia* using `canaper` is simple, requiring only two commands:
 
@@
 #| eval: false
-library(canaper)
 
-# Run randomization test, returning output as tibble
+# 1. Run randomization test
 acacia_rand_res <- cpr_rand_test(
@@
 
-# Classify endemism
+# 2. Classify endemism
 acacia_canape <- cpr_classify_endem(acacia_rand_res)
@@
 Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
-Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr).
+Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr):
 
@@
 b <- ggplot(
-  acacia_canape,
-  aes(x = pe_alt_obs, y = pe_obs, color = endem_type)
-) +
+  acacia_canape, aes(x = pe_alt_obs, y = pe_obs, color = endem_type)) +
   geom_abline(slope = 1, color = "darkgrey") +
@@
 
-As in Figure 3 of @Mishler2014, grid cells with significant endemism are primarily located on the coasts, with mostly non-significant grid-cells in the interior (@fig-acacia-cpr).
+As in Figure 3 of @Mishler2014, grid-cells with significant endemism are primarily located on the coasts, with mostly non-significant grid-cells in the interior (@fig-acacia-cpr).
 Furthermore, the endemism types largely correspond between the two figures.
@@
 
+## Comparison with Biodiverse
+
 We also re-ran the *Acacia* analysis using Biodiverse with the same settings as @Mishler2014 (999 replicates of the `rand_structured` null model) and compared the results with those from `canaper`.
 Importantly, there is no expectation that results between the two should match exactly, for two reasons.
-First, the null model used between the Biodiverse analysis (`rand_structured`) and the `canaper` (`curveball`) differs.
+First, the null model used between the Biodiverse and `canaper` analyses differ (`rand_structured` and `curveball`, respectively).
 `rand_structured` is not currently available in R, but we hope to add this to a future version of `canaper`.
 Second, the random communities generated in each run will be different, so the exact *p*-values will also be different.
-With a sufficiently high number of random communities, significance (e.g., at the $\alpha = 0.5$ level) is expected to converge, but there may be borderline cases that appear signfificant in some analyses and non-significant in others.
+With a sufficiently high number of random communities, significance (e.g., at the $\alpha = 0.5$ level) is expected to converge, but there may be borderline cases that appear significant in some analyses and non-significant in others.
 
 When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells.
-Differences between the two are mostly due to a slightly greater number of significant sites in the `canaper` results (@tbl-cpr-biod).
+Differences between the two are mostly due to a greater number of significant sites in the `canaper` results (@tbl-cpr-biod).
 Again, this does not mean that one or the other is more correct; rather, the differences in significant endemism likely reflect differences in the null models that produced the random communities.
-Here, both the `rand_structured` and `curveball` null models preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models, which may explain the relative high agreement between results.
+Here, both the `rand_structured` and `curveball` null models preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models [@Strona2018], which may explain the relative high agreement between results.
 Selection of an appropriate null model is beyond the scope of this paper, but must be considered carefully in any community ecological analysis.
@@
 Some of those that are more closely related to `canaper` include the following.
-`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD.
+`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000].
 `phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing speed of PD, and is used by `canaper`.
@@
 The `canaper` package enables CANAPE completely within R for the first time.
-This will greatly simplify workflows and facilitate reproducibility for researchers studying spatial biodiversity with R.
+This will simplify workflows and facilitate reproducibility for researchers studying spatial biodiversity with R.
 Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
@@
 
-(a) Map of Australia showing grid cells (communities) colored by endemism type.
+\(a) Map of Australia showing grid-cells (communities) colored by endemism type.
 (b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.
@@
 # Tables
+
+```{r}
+#| label: tbl-datasets
+#| tbl-cap: "Datasets in `canaper`"
+#| echo: false
+
+comms <- list(
+  acacia = acacia$comm,
+  biod_example = biod_example$comm,
+  phylocom = phylocom$comm
+)
+
+phylos <- list(
+  acacia = acacia$phy,
+  biod_example = biod_example$phy,
+  phylocom = phylocom$phy
+)
+
+tibble(
+  dataset = names(comms),
+  n_species_comm = map_dbl(comms, ncol),
+  n_sites = map_dbl(comms, nrow),
+  n_species_tree = map_dbl(phylos, ape::Ntip)
+) %>%
+  knitr::kable()
+```
+
+The `acacia` dataset [@Mishler2014] includes two outgroup taxa in the phylogeny that are not in the community matrix.
+The `phylocom` dataset [@Webb2008] also includes more taxa in the phylogeny than the community matrix.
 
@@
 
-**Box 1**: Columns in dataframe produced by `cpr_rand_test()`.
+**Box 1**: Glossary of terms.
 
-`cpr_rand_test()` calculates four basic metrics: `pd`, phylogenetic diversity (PD) [@Faith1992]; `rpd` [@Mishler2014], relative PD; `pe` [@Rosauer2009], phylogenetic endemism (PE); `rpe` relative PE [@Mishler2014].
+- Community: A co-occurring set of species.
+- Grid-cell: One of a set of spatial units each with equal area that together cover a region of interest; each grid-cell comprises a community. Often used synonymously with "site".
+- Community matrix: A 2-dimensional dataset with communties (grid-cells) on one axis and species on the other. `canaper` expects communities as rows and species as columns.
+- Phylogenetic diversity (PD): The total branch length (including the root) of a phylogenetic tree that includes all species occurring in a given area (community or grid-cell) [@Faith1992].
+- Phylogenetic endemism (PE): PD inversely weighted by the distribution range (i.e., number of grid-cells) of a branch [@Rosauer2009].
+- Relative phylogenetic diversity (RPD): The ratio of PD measured on the original tree vs. PD measured on a comparison tree with all branch lengths transformed to equal length [@Mishler2014].
+- Relative phylogenetic endemism (RPE): The ratio of PE measured on the original tree vs. PE measured on a comparison tree with all branch lengths transformed to equal length [@Mishler2014].
+- Neoendemic: An area with a high concentration of short, range-restricted branchlengths; may be due to processes such as recent speciation (radiation).
+- Paleoendemic: An area with a high concentration of long, range-restricted branchlengths; may be due to processes such as extinction or colonization by distantly related species.
+- Mixed endemic: An area with a mixture of short and long, range-restricted branchlengths; may be due to multiple processes.
+- Super endemic: An area with mixed endemism that is statistically highly significant.
 
-Four each of the four basic metrics, 9 columns are calculated:
+**Box 2**: Columns in dataframe produced by `cpr_rand_test()`.
 
-- `*_obs`: Observed value
-- `*_obs_c_lower`: Count of times observed value was lower than random values
-- `*_obs_c_upper`: Count of times observed value was higher than random values
-- `*_obs_p_lower`: Percentage of times observed value was lower than random values
-- `*_obs_p_upper`: Percentage of times observed value was higher than random values
-- `*_obs_q`: Count of the non-NA random values used for comparison
-- `*_obs_z`: Standard effect size (*z*-score)
-- `*_rand_mean`: Mean of the random values
-- `*_rand_sd`: Standard deviation of the random values
+`cpr_rand_test()` calculates six basic metrics for each grid-cell (Box 1):
+
+- `pd`: phylogenetic diversity (PD).
+- `pd_alt`: alternative PD (PD measured on comparison tree with all branch lengths transformed to equal length).
+- `pe`: phylogenetic endemism (PE).
+- `pe_alt`: alternative PE (PE measured on comparison tree).
+- `rpd`: relative PD (ratio of `pd` to `pd_alt`).
+- `rpe`: relative PE (ratio of `pe` to `pe_alt`).
+
+For each of the six basic metrics, nine columns (54 total) are output by `cpr_rand_test()`:
+
+- `*_obs`: Observed value.
+- `*_obs_c_lower`: Count of times observed value was lower than random values.
+- `*_obs_c_upper`: Count of times observed value was higher than random values.
+- `*_obs_p_lower`: Percentage of times observed value was lower than random values.
+- `*_obs_p_upper`: Percentage of times observed value was higher than random values.
+- `*_obs_q`: Count of the non-NA random values used for comparison.
+- `*_obs_z`: Standard effect size (*z*-score).
+- `*_rand_mean`: Mean of the random values.
+- `*_rand_sd`: Standard deviation of the random values.
 
 So for `pd` the output columns would include `pd_obs`, `pd_obs_c_lower`, etc.
+The *p*-values (`*_obs_p_upper`, etc.) are then used to categorize endemism type according to CANAPE (Box 3).
+
+**Box 3**: Criteria for categorical analysis of neo- and paleo-endemism (CANAPE).
+
+Significance assessed at $\alpha = 0.05$ (significant) or $\alpha = 0.01$ (highly significant).
+Metrics described in Box 2.
+
+- If either `pe_obs_p_upper` or `pe_alt_obs_p_upper` are significantly high (one-sided test), then look for paleo- or neo-endemism.
+  - If `rpe_obs_p_upper` is significantly high (two-sided test), then grid-cell is paleo-endemic.
+  - Else if `rpe_obs_p_lower` is significantly low (two-sided test), then grid-cell is neo-endemic.
+  - Else there is mixed age endemism, in which case:
+    - If both `pe_obs_p_upper` and `pe_alt_obs_p_upper` are highly significant (*p* < 0.01), then grid-cell is super-endemic.
+    - Else grid-cell is mixed-endemic.
+- Else if neither `pe_obs_p_upper` nor `pe_alt_obs_p_upper` are significantly high, grid-cell is non-endemic.

### 2022-09-21 ms/ms.Qmd
Update MS
format structure for MEE

@@
 Randomization algorithms are provided via the `vegan` package [@Oksanen2017], which comes with \> 30 pre-defined algorithms as well as the option to provide a custom, user-defined algorithm.
-`canaper` has undergone thorough code review through ROpenSci [@Boettiger2015] and is verified against a large number of unit tests (>98% coverage).
+`canaper` has undergone thorough code review through ROpenSci [@Boettiger2015] and is verified against a large number of unit tests (>95% coverage).
 All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.
@@
 Community data may be either presence-absence data (0s or 1s) or abundance data (integers \>= 0).
-However, all calculations of PD and PE only use presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
+However, when PD and PE are calculated only presence-absence information is used (i.e., abundance weighting is not used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
 Community data is typically loaded using `read.csv()`, `readr::read_csv()`, or other functions that can import rectangular data.
@@
 
-# Conclusion
+# Conclusions
 
@@
 
+# Acknowledgements
+
+This research supported in part by Japan Society for the Promotion of Science (Kakenhi) Grant numbers 16H06279 and 22H04925.
+Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
+
+# Conflict of interest
+
+All authors declare no conflict of interest.
+
+# Author contributions
+
+Joel H. Nitta conceived the ideas and designed methodology; Joel H. Nitta and Shawn W. Laffan analysed the data; Joel H. Nitta led the writing of the manuscript.
+All authors contributed critically to the drafts and gave final approval for publication.
+
+# Data availability
+
+All data and code are available in the `canaper` R package, which is distributed by the Comprehensive R Archive Network (CRAN).
+Source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
+Usage of `canaper` is documented at https://docs.ropensci.org/canaper/.
+The `acacia` dataset was originally published by @Mishler2014.
+
 # References
@@
 
-# Figures
-```{r}
-#| label: fig-acacia-cpr
-#| fig.height: 8
-#| fig-cap: "Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*."
-#| echo: false
+{{< pagebreak >}}
 
-# Fist do some data wrangling to make the results easier to plot
-# (add lat/long columns)
-acacia_canape <- acacia_canape |>
-  separate(site, c("long", "lat"), sep = ":") |>
-  mutate(across(c(long, lat), parse_number))
+# Figure captions
 
-a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
-  geom_tile() +
-  # cpr_endem_cols is a color palette in canaper for endemism colors
-  scale_fill_manual(values = cpr_endem_cols) +
-  guides(
-    fill = guide_legend(title.position = "top", label.position = "bottom")
-  ) +
-  coord_fixed() +
-  theme(legend.position = "bottom", legend.title = element_blank())
-
-b <- ggplot(
-  acacia_canape,
-  aes(x = pe_alt_obs, y = pe_obs, color = endem_type)
-) +
-  geom_abline(slope = 1, color = "darkgrey") +
-  geom_point() +
-  scale_color_manual(values = cpr_endem_cols) +
-  labs(
-    x = "Phylogenetic endemism on comparison tree",
-    y = "Phylogenetic endemism on actual tree"
-  ) +
-  theme_bw() +
-  theme(legend.position = "none")
-
-a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
-```
-
-\(a) Map of Australia showing grid-cells (communities) colored by endemism type.
+**@fig-acacia-cpr.** Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*.
+(a) Map of Australia showing grid-cells (communities) colored by endemism type.
 (b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.
@@
 This figure reproduces Figure 3 of @Mishler2014.
+
+{{< pagebreak >}}
 
@@
 ```
+
+{{< pagebreak >}}
 
@@
 - Else if neither `pe_obs_p_upper` nor `pe_alt_obs_p_upper` are significantly high, grid-cell is non-endemic.
+
+```{r}
+#| label: fig-acacia-cpr
+#| fig.height: 8
+#| fig-cap: ""
+#| echo: false
+
+# Fist do some data wrangling to make the results easier to plot
+# (add lat/long columns)
+acacia_canape <- acacia_canape |>
+  separate(site, c("long", "lat"), sep = ":") |>
+  mutate(across(c(long, lat), parse_number))
+
+a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
+  geom_tile() +
+  # cpr_endem_cols is a color palette in canaper for endemism colors
+  scale_fill_manual(values = cpr_endem_cols) +
+  guides(
+    fill = guide_legend(title.position = "top", label.position = "bottom")
+  ) +
+  coord_fixed() +
+  theme(legend.position = "bottom", legend.title = element_blank())
+
+b <- ggplot(
+  acacia_canape,
+  aes(x = pe_alt_obs, y = pe_obs, color = endem_type)
+) +
+  geom_abline(slope = 1, color = "darkgrey") +
+  geom_point() +
+  scale_color_manual(values = cpr_endem_cols) +
+  labs(
+    x = "Phylogenetic endemism on comparison tree",
+    y = "Phylogenetic endemism on actual tree"
+  ) +
+  theme_bw() +
+  theme(legend.position = "none")
+
+a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
+```

### 2022-09-22 ms/ms.Qmd
Update ms
in response to comments from BM, SL 2022-09-20

@@
 
-Joel H. Nitta, Shawn W. Laffan, Brent D. Mishler, and Wataru Iwasaki
+Joel H. Nitta^1,\*^, Shawn W. Laffan^2^, Brent D. Mishler^3^, and Wataru Iwasaki^1,4,5,6,7^
+
+^1^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
+
+^2^Department of Integrative Biology, University of California Berkeley, Berkeley, USA
+
+^3^School of Biological, Earth and Environmental Sciences, The University of New South Wales, Kensington, Australia
+
+^4^Department of Computational Biology and Medical Sciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
+
+^5^Atmosphere and Ocean Research Institute, The University of Tokyo, Chiba, Japan
+
+^6^Institute for Quantitative Biosciences, The University of Tokyo, Tokyo, Japan
+
+^7^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan
+
+\* Corresponding author  
+Joel H. Nitta  
+Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan  
+joelnitta@gmail.com
+
+Running headline  
+*canaper*
 
@@
 
-2. Here, we describe `canaper`, a new R package that provides functions to conduct CANAPE in R.
+2. Here, we present `canaper`, a new R package that provides functions to conduct CANAPE in R.
 `canaper` implements methods for efficient computation, including parallelization and encoding of community data as sparse matrices.
@@
 Until recently, quantification of biology was done on a purely taxonomic basis: counting the number of species in an area, for example [e.g., @Diamond1975].
-However, as all taxa are related to some degree by descent from a common ancestor, a thorough understanding of biodiversity is only possible by taking into account their evolutionary relationships.
-This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity [PD\; @Faith1992] and phylogenetic endemism [PE\; @Faith1992; @Rosauer2009] (Box 1).
-Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies [e.g., @Mishler2020; @Nitta2022a; @Thornhill2017].
+However, as all taxa are related to some degree by descent from a common ancestor, a thorough understanding of biodiversity is only possible by considering their evolutionary relationships.
+This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity [PD\; @Faith1992] and phylogenetic endemism [PE\; @Rosauer2009] (Box 1).
+Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies and large spatial datasets [e.g., @Mishler2020; @Nitta2022a; @Thornhill2017].
 
 One recently developed extension of PE is categorical analysis of neo-and paleo-endemism [CANAPE\; @Mishler2014].
-CANAPE uses phylogenetic methods to infer the evolutionary processes that give rise to certain patterns of endemism.
-In theory, endemic areas (areas with high rates of range-limited taxa; or in the phylogenetic sense, range-limited portions of the phylogenetic tree) may arise via two major processes: previously wide-spread lineages may undergo extinction in all but a portion of their range (paleo-endemism), or recently diverged lineages may only occur in a small area (neo-endemism).
-It is also possible that a given area is home to both paleo- and neo-endemic lineages (mixed endemism).
-CANAPE involves analyzing observed patterns of phylogentic endemism in comparison with a null model to infer whether areas are paleo-endemic, neo-endemic, mixed, or lack significantly high endemism.
-CANAPE has been used in \> 50 publications (Google Scholar query for papers that cite @Mishler2014 and mention "CANAPE"), and is a central component of the field of spatial phylogenetics.
+CANAPE uses phylogenetic methods to give insight into evolutionary processes that produce endemism.
+In theory, centers of endemism (areas with many range-limited taxa; or in the phylogenetic sense, concentrations of range-limited branches of the phylogeny) may arise via multiple processes.
+For example, previously widespread lineages may undergo extinction in all but a portion of their range (paleo-endemism), or recently diverged lineages may only occur in a small area (neo-endemism). 
+It is also possible that a given area is home to a high concentration of both paleo- and neo-endemic lineages (mixed endemism).
+CANAPE involves analyzing observed patterns of PE in comparison with a null model to infer first whether an area is significantly high in PE, then classifying those areas into centers of paleo-endemism, neo-endemism, or mixed endemism.
+CANAPE is widely used (\> 50 publications in Google Scholar query for papers that cite @Mishler2014 and mention "CANAPE"), and is a central component of the field of spatial phylogenetics.
 
-Despite the popularity of CANAPE, it has so far only been implemented in one piece of software, Biodiverse [@Laffan2010].
-Biodiverse is written in perl and includes a graphical user interface (GUI).
+Despite the popularity of CANAPE, it has so far only been implemented in one software package, Biodiverse [@Laffan2010].
+Biodiverse is written in perl and comprises an analytical engine and a graphical user interface (GUI).
 While Biodiverse is convenient for non-coders because of its GUI, many ecologists and evolutionary biologists use R for their analyses [@Lai2019].
-R is well-suited for reproducible workflows and is extremely versatile thanks to the thousands of community-contributed packages that extend its utility.
-Until now, an R user who wanted to conduct CANAPE analysis as part of a broader workflow needed to first clean raw data in R, export it to Biodiverse, conduct PD and PE analyses in Biodiverse, then import the results back into R for further analysis and visualization.
-Also, while Biodiverse conducted the metrics needed to categorize endemism type, it did not actually perform this categorization.
-Besides being error-prone due to its manual nature, this workflow is also not suited to parallel processing, which is needed for large datasets.
-Previously, parallel processing and automation of such analyses was only possible using perl scripts; a set of R scripts is available to call perl from R (<https://github.com/NunzioKnerr/biodiverse_pipeline>), but it is not an R package and is not straightforward to use.
+Until now, an R user who wanted to conduct CANAPE analysis as part of a broader R workflow needed to first clean raw data, export it to Biodiverse, conduct PD and PE analyses in Biodiverse, then import the results back into R for further analysis and visualization.
+Also, while Biodiverse conducted the metrics needed to categorize endemism type, it did not actually perform this categorization (although this feature is planned to be added to a future version of Biodiverse).
+Furthermore, the Biodiverse GUI alone does not support parallel processing, which is needed for large datasets.
+Parallel processing and automation of Biodiverse analyses is only possible using perl scripts.
+A set of R scripts is available to call Biodiverse perl scripts from R (<https://github.com/NunzioKnerr/biodiverse_pipeline>), but it is not an R package and does not conduct CANAPE within R.
 
-Here, we describe a new R package that implements CANAPE completely in R, `canaper`.
+Here, we present a new R package that implements CANAPE completely in R: `canaper`.
 We strived to make `canaper` simple to use and efficient.
 Parallel computing can be enabled with a single line of code.
-Randomization algorithms are provided via the `vegan` package [@Oksanen2017], which comes with \> 30 pre-defined algorithms as well as the option to provide a custom, user-defined algorithm.
 `canaper` has undergone thorough code review through ROpenSci [@Boettiger2015] and is verified against a large number of unit tests (>95% coverage).
@@
 Community data may be either presence-absence data (0s or 1s) or abundance data (integers \>= 0).
-However, when PD and PE are calculated only presence-absence information is used (i.e., abundance weighting is not used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
+However, all calculations of PD and PE use only presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
 Community data is typically loaded using `read.csv()`, `readr::read_csv()`, or other functions that can import rectangular data.
@@
 First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (`pd_obs`) and phylogenetic endemism (`pe_obs`) are calculated.
-Next, an alternative phylogeny (also referred to as the "comparison tree") is constructed that has non-zero branch lengths each set to an equal value, then rescaled to a total length of 1.
+Next, an alternative phylogeny (also referred to as the "comparison tree") is constructed that has non-zero branch lengths set to a constant value, then rescaled to a total length of 1.
 PD and PE are then calculated on the alternative phylogeny (`pd_alt`, `pe_alt`).
 Relative PD and PE, the ratio of `pd_obs` to `pd_alt` and `pe_obs` to `pe_alt` respectively, are then calculated (RPD, RPE).
-In CANAPE, areas (grid-cells) with extremely high or low RPE are designated as paleo- or neo-endemic, respectively.
-However, the statistical significance of any given RPE value cannot be determined from observed values alone.
+However, the statistical significance of any of these metrics cannot be determined from observed values alone.
 
@@
 
-PD (and by extension, PE, RPD, and RPE) is expected to increase with taxon richness, since the chance of drawing two distantly related taxa without replacement decreases as richness increases.
-To determine the statistical significance of RPE, the observed value is compared to a distribution of values obtained from a set of random communities that each have the same richness as the original data.
+PD (and by extension, PE, RPD, and RPE) will increase with taxon richness, since adding a new taxon adds new branches to the tree.
+To determine the statistical significance of these metrics, the observed value is compared to a distribution of values obtained from a set of random communities.
 The random communities are generated by a randomization algorithm that shuffles the original data.
 
-Since the randomization algorithm determines the range of reference (expected) values, the choice of randomization algorithm is likely to have a large effect on the results.
+Since the randomization algorithm influences the range of reference (expected) values, the choice of randomization algorithm is likely to have a large effect on the results.
 As there is no single "correct" algorithm, we have opted to provide the user with a wide range of options by implementing randomization algorithms included in the `vegan` package.
@@
 Recommended algorithms include `swap` [@Gotelli2003] and `curveball` [@Strona2014].
-These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to the randomization algorithm in Biodiverse, `rand_structured` (see "Example: Australian *Acacia*").
+These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to one commonly used randomization algorithm in Biodiverse called `rand_structured` (see "Example: Australian *Acacia*").
 
@@
 Once a randomization algorithm has been selected, random communities are generated for a number of replicates set by the user, and a set of summary statistics are computed (Box 2).
-Summary statistics include the mean and standard deviation of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the random communities including standard effect size and rank, which is used to calculate *p*-value.
+Summary statistics include the mean and standard deviation of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the random communities including standard effect size and rank, which is used to calculate *p*-values.
 
-To maximize computing efficiency, the individual random communities are not stored in memory.
-Since 100s to 1000s or more random replicates may be required, the memory usage can be quite high for large matrices.
+To minimize memory requirements, the individual random communities are not stored in memory.
+Since 100s to 1000s or more random replicates may be required, the memory usage could be quite high for large matrices if each replicate was stored.
 Rather, as each random community is generated, the summary statistics are computed immediately and stored.
-This way, for any input matrix of size $N \times M$ and $n$ random replicates, where $N$ is number of rows (sites) and $M$ is number columns (species), $n$ summary matrices each with $N$ rows and one column per summary statistic are stored, rather than $n$ matrices of size $N \times M$.
+This way, for any input matrix of size $N \times M$ and $n$ random replicates, where $N$ is number of rows (sites) and $M$ is number columns (species), $n$ summary matrices (dataframes) each with $N$ rows and one column per summary statistic are stored, rather than $n$ matrices of size $N \times M$ (the same approach is used in Biodiverse).
 
@@
 Briefly, this is done by comparing significance values of summary statistics calculated in the previous step (Box 3).
-In order to be considered significantly endemic, a given grid-cell must first have significantly high `pe_obs` or `pe_alt` or both.
-If this is true, the grid-cell is classified into one of three non-overlapping categories: if the grid-cell has significantly high or low RPE, it is considered to be a center of paleo-endemism or neo-endemism, respectively; if RPE is not signficant (but `pe_obs`, `pe_alt`, or both are), it is considered a center of mixed endemism.
-Centers of mixed endemism are futher divided based on *p*-value; if `pe_obs` and `pe_alt` and both significant at the $\alpha = 0.01$ level, the grid-cell is considered a center of super-endemism.
+To be considered significantly endemic, a given grid-cell must first have significantly high `pe_obs` or `pe_alt` or both (one-tailed test).
+If this is true, the grid-cell is classified into one of three non-overlapping categories: if the grid-cell has significantly high or low RPE (two-tailed test), it is considered to be a center of paleo-endemism or neo-endemism, respectively; if RPE is not significant (but `pe_obs`, `pe_alt`, or both are), it is considered a center of mixed endemism.
+Centers of mixed endemism can be further divided based on *p*-value; if `pe_obs` and `pe_alt` and both significant at the $\alpha = 0.01$ level, the grid-cell may be considered a center of super-endemism (but not all studies make this distinction).
 
@@
 Binary algorithms are designed for binary (i.e., presence-absence) data, and quantitative algorithms are designed for quantitative (i.e., abundance) data.
-Either type of algorithm will accept either type of data, but binary algorithms will convert abundance data to binary and return a binary matrix.
+Either type of algorithm will accept either type of data, but binary algorithms will convert abundance data to binary and return a binary matrix (dataframe).
 
@@
 The `cpr_rand_test()` function carries out calculation of observed values, generation of random communities, and calculation of summary statistics as described above in "Analysis workflow".
-The main arguments to this function are the input community and phylogeny, type of null model, and various settings for the null model.
+The main arguments to this function are the input community and phylogeny, type of null model, and settings for the null model.
 For a full list of null models to choose from, run `?vegan::commsim()`.
-It should be noted that the type of null model, number of random communities, and number of swapping iterations performed per random community (for swapping algorithms) all may strongly affect results of `cpr_rand_test()` [or any metric that is based on comparison to a set of random communties\; @Gotelli2001a]
-While it is beyond the scope of this paper to provide a full discussion of null models in ecology, we have provided details about how to explore appropriate null model settings with `canaper` in the "How many randomizations?" vignette (<https://joelnitta.github.io/canaper/articles/how-many-rand.html>).
+It should be noted that the type of null model, number of random communities, and number of swapping iterations performed per random community (for swapping algorithms) all may strongly affect results of `cpr_rand_test()` [or any metric that is based on comparison to a set of random communities\; @Gotelli2001a].
+While it is beyond the scope of this paper to provide a full discussion of null models in ecology, we have provided details about how to explore appropriate null model settings with `canaper` in the "How many randomizations?" vignette (<https://docs.ropensci.org/canaper/articles/how-many-rand.html>).
 
@@
 
-# Look a subset of the output columns
+# Print a subset of the output columns
 rand_test_results[, 1:9]
@@
 The `cpr_classify_endem()` function classifies endemism types for the output of `cpr_rand_test()` as described above in "Analysis workflow".
-The input is a dataframe including the following columns calculated by `cpr_rand_test()`: `pe_obs_p_upper` (upper *p*-value comparing observed phylogenetic endemism to random values), `pe_alt_obs_p_upper` (upper *p*-value comparing observed phylogenetic endemism on alternate tree to random values), and `rpe_obs_p_upper` (upper *p*-value comparing observed relative phylogenetic endemism to random values).
+The input is a dataframe including the following columns calculated by `cpr_rand_test()`: `pe_obs_p_upper` (upper *p*-value comparing observed PE to random values), `pe_alt_obs_p_upper` (upper *p*-value comparing observed PE on alternate tree to random values), `rpe_obs_p_upper` (upper *p*-value for RPE), and `rpe_obs_p_lower` (lower *p*-value for RPE).
 The output is the same dataframe, with the column `endem_type` appended.
@@
 `picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000].
-`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing speed of PD, and is used by `canaper`.
+`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing efficiency of PD, and is used by `canaper`.
 `vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
-`vegan` includes by far the greatest number of algorithms for generating random communities (> 30), and is used by `canaper`.
+`vegan` includes by far the greatest number of algorithms for generating random communities (> 30) and is used by `canaper`.
 `adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.
-`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics, but does not appear to be currently maintained and has been removed from CRAN.
+`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics but does not appear to be currently maintained and has been removed from CRAN.
 
@@
 Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
+Luis Osorio and Klaus Schliep reviewed the code as part of submission to ROpenSci.
 
@@
 
-Joel H. Nitta conceived the ideas and designed methodology; Joel H. Nitta and Shawn W. Laffan analysed the data; Joel H. Nitta led the writing of the manuscript.
+Joel H. Nitta conceived the ideas and designed methodology; Joel H. Nitta and Shawn W. Laffan analyzed the data; Joel H. Nitta led the writing of the manuscript.
 All authors contributed critically to the drafts and gave final approval for publication.
@@
 tibble(
-  dataset = names(comms),
-  n_species_comm = map_dbl(comms, ncol),
-  n_sites = map_dbl(comms, nrow),
-  n_species_tree = map_dbl(phylos, ape::Ntip)
+  Dataset = names(comms),
+  `*n* sites` = map_dbl(comms, nrow),
+  `*n* species (community matrix)` = map_dbl(comms, ncol),
+  `*n* species (tree)` = map_dbl(phylos, ape::Ntip)
 ) %>%
@@
 
-knitr::kable(comp_table)
+comp_table %>%
+  select(
+    `Endemism type` = endem_type,
+    `*n* sites (canaper)` = n_canaper,
+    `*n* sites (Biodiverse)` = n_biodiverse
+  ) %>%
+knitr::kable()
 ```
@@
 Metrics described in Box 2.
+Not all studies include super-endemism.
 
@@
 #| fig.height: 8
-#| fig-cap: ""
+#| fig-cap: "Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*."
 #| echo: false

### 2022-09-22 _targets.R
Use biodiverse results run on MacBook pro

@@ -44,7 +44,7 @@ tar_plan(
   # Load results from Biodiverse for Acacia
   tar_file_read(
     acacia_biod_res,
-    "_targets/user/data_raw/CANAPE_Acacia/spatial_results_prank.csv",
+    "_targets/user/data_raw/acacia_prank.csv",
     read_csv(!!.x)),
   # Classify endemism types for Biodiverse results
   acacia_canape_biod = classify_biod(acacia_biod_res),

### 2022-09-22 R/functions.R
Add check for hardware specs before running canape

@@ -1,11 +1,36 @@
+# Get hardware specs for a mac
+get_mac_hardware_info <- function() {
+  system("system_profiler SPHardwareDataType", intern = TRUE)
+}
+
+#' Wrapper for canaper::cpr_rand_test()
+#'
+#' Additional arguments:
+#' @param workers Number of CPUs to use in parallel
+#' @param hardware_info Character vector; output of get_mac_hardware_info(),
+#' includes hardware specs when running on a Mac
+#' @param req_spec Character vector; strings to detect in `hardware_info`;
+#' if any are not detected, the function will fail
 run_canape <- function(
-  comm, phy, null_model, n_reps = 500, n_iterations = 100000, workers = 6) {
+  comm, phy, null_model, n_reps = 500, n_iterations = 100000, workers = 6,
+  hardware_info = NULL, req_spec = NULL) {
   on.exit(plan(sequential))
-  # Specify two cores running in parallel
+  # Check hardware requirements
+  if (!is.null(hardware_info)) {
+    hardware_check <- purrr::map_lgl(
+      req_spec,
+      ~any(stringr::str_detect(hardware_info, fixed(.)))
+    )
+    assertthat::assert_that(
+      all(hardware_check),
+      msg = "Required hardware not detected"
+    )
+  }
+  # Set up parallelization
   plan(multisession, workers = workers)
   # Run randomization test, returning output as tibble (not a dataframe)
   rand_res <- cpr_rand_test(
-    comm, phy,
+    comm = comm, phy = phy,
     null_model = null_model,
     n_reps = n_reps, n_iterations = n_iterations,
     tbl_out = TRUE

### 2022-09-22 _targets.R

@@ -22,7 +22,15 @@ tar_plan(
     null_model = "curveball",
     n_reps = 999,
     n_iterations = 100000,
-    workers = 2
+    workers = 2,
+    hardware_info = hardware_specs,
+    # Require this be run on a mac laptop
+    hardware_specs = get_mac_hardware_info(),
+    req_spec = c(
+      "Model Name: MacBook Pro",
+      "Processor Name: Quad-Core Intel Core i7",
+      "Total Number of Cores: 4",
+      "Memory: 16 GB")
   ),
   # Repeat, with only one core for comparison
   acacia_canape_cpr_3 = run_canape(

### 2022-09-22 ms/ms.Qmd
Run canape on laptop

@@
 
-# Enforce dependency on acacia_canape_cpr_single
-tar_load(acacia_canape_cpr_single,
+# Enforce dependency on acacia_canape_cpr_1
+tar_load(acacia_canape_cpr_1,
   store = here::here("_targets"))
@@
 Computations were carried out on an iMac (2019) with 16 GB RAM and a 3GHz, six core processor.
-Approximate compute times were 1.3 hr for Biodiverse and `r times$acacia_canape_cpr_single$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
+Approximate compute times were 1.3 hr for Biodiverse and `r times$acacia_canape_cpr_1$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
 However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr$minutes` min (further increases in number of cores did not significantly decrease time).

### 2022-09-22 _targets.R

@@ -15,7 +15,12 @@ tar_plan(
     n_iter = magrittr::raise_to_power(10, 1:7),
     binary = TRUE
   ),
-  # Run CANAPE using canaper (2 cores)
+  # Run CANAPE using canaper (2 cores) on mac laptop
+  mac_laptop_specs = c(
+    "Model Name: MacBook Pro",
+    "Processor Name: Quad-Core Intel Core i7",
+    "Total Number of Cores: 4",
+    "Memory: 16 GB"),
   acacia_canape_cpr = run_canape(
     comm = acacia$comm,
     phy = acacia$phy,
@@ -23,31 +28,30 @@ tar_plan(
     n_reps = 999,
     n_iterations = 100000,
     workers = 2,
-    hardware_info = hardware_specs,
-    # Require this be run on a mac laptop
-    hardware_specs = get_mac_hardware_info(),
-    req_spec = c(
-      "Model Name: MacBook Pro",
-      "Processor Name: Quad-Core Intel Core i7",
-      "Total Number of Cores: 4",
-      "Memory: 16 GB")
+    hardware_info = get_mac_hardware_info(),
+    req_spec = mac_laptop_specs
   ),
-  # Repeat, with only one core for comparison
+  # Repeat, with three cores for comparison
   acacia_canape_cpr_3 = run_canape(
     comm = acacia$comm,
     phy = acacia$phy,
     null_model = "curveball",
     n_reps = 999,
     n_iterations = 100000,
-    workers = 3
+    workers = 3,
+    hardware_info = get_mac_hardware_info(),
+    req_spec = mac_laptop_specs
   ),
-  acacia_canape_cpr_single = run_canape(
+  # Repeat, with one core for comparison
+  acacia_canape_cpr_1 = run_canape(
     comm = acacia$comm,
     phy = acacia$phy,
     null_model = "curveball",
     n_reps = 999,
     n_iterations = 100000,
-    workers = 1
+    workers = 1,
+    hardware_info = get_mac_hardware_info(),
+    req_spec = mac_laptop_specs
   ),
   # Load results from Biodiverse for Acacia
   tar_file_read(

### 2022-09-22 ms/ms.Qmd
Update MS

@@
 highlight-style: github
+crossref: 
+  fig-title: Fig.
+  tbl-title: Table
 format:
@@
 To demonstrate usage with a real dataset, we reproduced the analysis of @Mishler2014, who conducted CANAPE on Australian *Acacia* using Biodiverse.
+All `canaper` analyses were run with `canaper` v`r packageVersion("canaper")` in R v`r R.Version()$major`.`r R.Version()$minor`.
 Conducting CANAPE on Australian *Acacia* using `canaper` is simple, requiring only two commands:
@@
 
-We also re-ran the *Acacia* analysis using Biodiverse with the same settings as @Mishler2014 (999 replicates of the `rand_structured` null model) and compared the results with those from `canaper`.
+We also re-ran the *Acacia* analysis using Biodiverse v3.1 with the same settings as @Mishler2014 (999 replicates of the `rand_structured` null model) and compared the results with those from `canaper`.
 Importantly, there is no expectation that results between the two should match exactly, for two reasons.
@@
 
-Computations were carried out on an iMac (2019) with 16 GB RAM and a 3GHz, six core processor.
+Computations were carried out on an MacBook Pro (2019) with 16 GB RAM and a 1.7 GHz, four core processor.
 Approximate compute times were 1.3 hr for Biodiverse and `r times$acacia_canape_cpr_1$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
 However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr$minutes` min (further increases in number of cores did not significantly decrease time).
-This demonstrates that `canaper` can efficiently compute moderately sized datasets with a personal computer.
+This demonstrates that `canaper` can efficiently compute moderately sized datasets with a personal laptop computer.
 
@@
 
-a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
+fig <- a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
+
+ggsave(plot = fig, file = "ms/figure_1.pdf", width = 100, height = 200, units = "mm")
+
+fig
 ```

### 2022-09-22 _targets.R
Add CANAPE using swap algorithm

@@ -53,6 +53,15 @@ tar_plan(
     hardware_info = get_mac_hardware_info(),
     req_spec = mac_laptop_specs
   ),
+  # Run CANAPE using independent swap algorithm
+  acacia_canape_cpr_swap = run_canape(
+    comm = acacia$comm,
+    phy = acacia$phy,
+    null_model = "swap",
+    n_reps = 999,
+    n_iterations = 10000,
+    workers = 2
+  ),
   # Load results from Biodiverse for Acacia
   tar_file_read(
     acacia_biod_res,

### 2022-09-28 ms/ms.Qmd
Update ms

@@
 The `cpr_rand_comm()` function generates a single random community. 
-The first argument, `comm` is a community data frame. 
+The first argument, `comm` is a community dataframe (or matrix). 
 The second, `null_model` is the name of one of the predefined randomization algorithms available in `vegan`.
@@
 
-The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with CANAPER, the test data from Phylocom [@Webb2008]:
+The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with `canaper`, the test data from Phylocom [@Webb2008]:
 
@@
 ```
+
+## `cpr_iter_sim()`
+
+```{r}
+#| label: acacia-stats
+#| include: false
+data(acacia)
+
+acacia_bin <- acacia$comm
+total_cells <- dim(acacia_bin)[1] * dim(acacia_bin)[2] 
+acacia_bin[acacia_bin > 0] <- 1
+acacia_non_empty <- 1 - (sum(acacia_bin) / total_cells)
+acacia_non_empty <- scales::percent(acacia_non_empty, accuracy = 0.1)
+```
+
+`cpr_iter_sim()` is not required for the CANAPE workflow, but is rather a diagnostic function for use with randomization algorithms that work by exchanging cells of the community matrix such as `swap` [@Gotelli2003] or `curveball` [@Strona2014].
+Each such exchange is known as an "iteration".
+It is generally impossible to know *a-priori* how many iterations are needed to completely randomize a given matrix; this number depends on properties of the matrix and the randomization algorithm.
+Generally, larger matrices with more skew (overabundance of zeros) will require more iterations (ADD CITATION).
+
+`cpr_iter_sim()` helps the user determine the minimum number of iterations needed by conducting successive exchanges and recording the similarity between the original matrix and the randomized matrix at each iteration.
+The similarity values should initially decrease until an approximate minimum is reached; further iterations will only result in noise around this minimum.
+A number slightly larger than the smallest number of iterations needed to reach the approximate minimum value can then be used for randomizing the community.
+The following code demonstrates usage of `cpr_rand_test()` with the *Acacia* dataset [@Mishler2014] that comes with `canaper`.
+The *Acacia* dataset is large (@tbl-datasets) and highly skewed (`r acacia_non_empty` zeros).
+
+```{r}
+#| label: cpr-iter-sim-demo
+data(acacia)
+
+# Conduct up to 100,000 iterations (swaps),
+# recording similarity every 1000 iterations
+iter_sim_res <- cpr_iter_sim(
+  comm = acacia$comm,
+  null_model = "curveball",
+  n_iterations = 100000,
+  thin = 1000
+)
+
+# Inspect the output
+iter_sim_res
+```
+
+It is useful to plot the output to identify the minimum number of iterations needed.
+Here we do so with the `ggplot2` package.
+
+```{r}
+#| label: cpr-iter-sim-plot
+#| fig.height: 8
+#| eval: false
+
+library(ggplot2)
+
+ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
+  geom_line() +
+  labs(x = "Num. iterations", y = "% Similarity")
+```
+
+The plot shows that the community matrix becomes maximally randomized after about 40,000--50,000 iterations (@fig-iter-sim), and that the original matrix and randomized matrix are about 96.5% similar.
 
@@
   geom_tile() +
-  # cpr_endem_cols is a color palette in canaper for endemism colors
-  scale_fill_manual(values = cpr_endem_cols) +
+  # use same colors as Misher et al. 2014
+  # (for CVD-safe colors see cpr_endem_cols or cpr_endem_cols_2)
+  scale_fill_manual(values = mishler_endem_cols) +
   guides(
@@
   geom_point() +
-  scale_color_manual(values = cpr_endem_cols) +
+  scale_color_manual(values = mishler_endem_cols) +
   labs(
@@
 
-The color palette used in @fig-acacia-cpr has been provided in `canaper` as `cpr_endem_cols`, and was selected to be distinguishable regardless of color vision type [@Okabe2002].
-We recommend using this color palette for maximum accessibility.
+The color palette used in @fig-acacia-cpr, `mishler_endem_cols`, is the same as that used in @Mishler2014.
+However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
+Alternate color palettes (e.g., `cpr_endem_cols`) are also vailable that include CVD-safe colors [@Okabe2002].
+Palettes for plotting CANAPE and *p*-rank results can be selected with the  `cpr_make_pal()` function.
 
 ## Comparison with Biodiverse
+
+```{r}
+#| label: tally-cells
+#| include: false
+
+# Tally number of significant cells in canper vs. Biodiverse
+sig_tally <-
+  comp_table %>%
+  mutate(
+    endem_type = str_replace_all(
+      endem_type, "not significant", "not_significant") %>%
+      as.factor() %>%
+      fct_collapse(significant = c("neo", "paleo", "mixed", "super"))
+      ) %>%
+  rename(canaper = n_canaper, biodiverse = n_biodiverse) %>%
+  pivot_longer(names_to = "analysis", values_to = "n_sites", -endem_type) %>%
+  group_by(endem_type, analysis) %>%
+  summarize(n_sites = sum(n_sites), .groups = "drop") %>%
+  split(.$analysis) %>%
+  purrr::map(., ~split(., .$endem_type))
+```
 
@@
 
-When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells.
-Differences between the two are mostly due to a greater number of significant sites in the `canaper` results (@tbl-cpr-biod).
-Again, this does not mean that one or the other is more correct; rather, the differences in significant endemism likely reflect differences in the null models that produced the random communities.
-Here, both the `rand_structured` and `curveball` null models preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models [@Strona2018], which may explain the relative high agreement between results.
+When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells (@tbl-cpr-biod).
+The total number of significant cells was nearly identical between the two (*n* = `r sig_tally$canaper$significant$n_sites` and `r sig_tally$biodiverse$significant$n_sites` sites, `canaper` and Biodiverse, respectively).
+The `rand_structured` and `curveball` null models both preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models [@Strona2018], which may explain the relative high agreement between results.
 Selection of an appropriate null model is beyond the scope of this paper, but must be considered carefully in any community ecological analysis.
@@
 #| include: false
+
+# Load times
+tar_load(acacia_biod_elapsed_min)
 
@@
 Computations were carried out on an MacBook Pro (2019) with 16 GB RAM and a 1.7 GHz, four core processor.
-Approximate compute times were 1.3 hr for Biodiverse and `r times$acacia_canape_cpr_1$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
+Approximate compute times were `r acacia_biod_elapsed_min` min for Biodiverse and `r times$acacia_canape_cpr_1$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
 However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr$minutes` min (further increases in number of cores did not significantly decrease time).
@@
 # Figure captions
+
+**@fig-iter-sim.** Testing the number of iterations needed to randomize the Australian *Acacia* dataset.
+The *Acacia* dataset [@Mishler2014] was randomized using the `curveball` algorithm for 100,000 iterations and percentage similarity between the original matrix and the randomized matrix was calculated every 1,000 iterations using the `cpr_iter_sim()` function.
 
@@
 ```{r}
+#| label: fig-iter-sim
+#| fig.height: 4
+#| fig-cap: "Testing the number of iterations needed to randomize the Australian *Acacia* dataset."
+#| echo: false
+
+fig <- ggplot2::ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
+  geom_line() +
+  labs(x = "Num. iterations", y = "% Similarity")
+
+ggsave(plot = fig, file = "ms/figure_1.pdf", width = 150, height = 150, units = "mm")
+
+fig
+```
+
+```{r}
 #| label: fig-acacia-cpr
@@
   geom_tile() +
-  # cpr_endem_cols is a color palette in canaper for endemism colors
-  scale_fill_manual(values = cpr_endem_cols) +
+  # use same colors as Misher et al. 2014
+  # (for CVD-safe colors could use cpr_endem_cols or cpr_endem_cols_2)
+  scale_fill_manual(values = mishler_endem_cols) +
   guides(
@@
 
-ggsave(plot = fig, file = "ms/figure_1.pdf", width = 100, height = 200, units = "mm")
+ggsave(plot = fig, file = "ms/figure_2.pdf", width = 100, height = 200, units = "mm")
 

### 2022-09-28 R/functions.R

@@ -79,29 +79,59 @@ test_iter <- function(comm, null_model, n_iter, binary) {
 #'
 #' PE_WE_P: Phylogenetic weighted endemism as a proportion of the total tree
 #' length
+#' = pe_obs_p_upper
 #'
 #' PHYLO_RPE2: Relative Phylogenetic Endemism (RPE). The ratio of the tree's PE
 #' to a null model where PE is calculated using a tree where all branches are of
 #' equal length.
 #' = rpe_obs_p_upper
 #'
-#' PHYLO_RPE_DIFF2: How much more or less PE is there than expected, in original
-#' tree units.
-#' = pe_obs_p_upper
-#'
-# PHYLO_RPE_NULL2: Null score used as the denominator in the RPE2 calculations
-# = pe_alt_obs_p_upper
-classify_biod <- function(biod_res) {
-  biod_res %>%
+#' PHYLO_RPE_NULL2: Null score used as the denominator in the RPE2 calculations
+#' = pe_alt_obs_p_upper
+#' @param biod_res_raw Output from Biodiverse (SPATIAL RESULTS, including
+#' only observed values)
+#' @param biod_res_prank Output from Biodiverse (p_rank >> SPATIAL RESULTS,
+#' including p-rank of observed values in comparison with randomizations)
+classify_biod <- function(biod_res_raw, biod_res_prank) {
+  biod_res_prank %>%
+  # Replace missing values with 0.5 as per SL code
+  # https://gist.github.com/shawnlaffan/2935dfd3efa92f23a440aa3db5916645
+  mutate(
+    PE_WE_P = case_when(
+      is.na(PE_WE_P) & (!is.na(PHYLO_RPE_NULL2) | !is.na(PHYLO_RPE2)) ~ 0.5,
+      TRUE ~ PE_WE_P
+    ),
+    PHYLO_RPE_NULL2 = case_when(
+      is.na(PHYLO_RPE_NULL2) & (!is.na(PE_WE_P) | !is.na(PHYLO_RPE2)) ~ 0.5,
+      TRUE ~ PHYLO_RPE_NULL2
+    ),
+    PHYLO_RPE2 = case_when(
+      is.na(PHYLO_RPE2) & (!is.na(PHYLO_RPE_NULL2) | !is.na(PE_WE_P)) ~ 0.5,
+      TRUE ~ PHYLO_RPE2
+    )
+  ) %>%
[…]

### 2022-09-28 _targets.R

@@ -59,21 +59,38 @@ tar_plan(
     phy = acacia$phy,
     null_model = "swap",
     n_reps = 999,
-    n_iterations = 10000,
-    workers = 2
+    n_iterations = 100000,
+    workers = 4
   ),
   # Load results from Biodiverse for Acacia
+  # - p-rank results
   tar_file_read(
-    acacia_biod_res,
-    "_targets/user/data_raw/acacia_prank.csv",
-    read_csv(!!.x)),
+    acacia_biod_res_prank,
+    "_targets/user/data_raw/acacia_canape_prank_spatial_results.csv",
+    read_csv(!!.x)
+  ),
+  # - raw (observed) values
+  tar_file_read(
+    acacia_biod_res_raw,
+    "_targets/user/data_raw/acacia_canape_spatial_results.csv",
+    read_csv(!!.x)
+  ),
   # Classify endemism types for Biodiverse results
-  acacia_canape_biod = classify_biod(acacia_biod_res),
+  acacia_canape_biod = classify_biod(
+    biod_res_raw = acacia_biod_res_raw,
+    biod_res_prank = acacia_biod_res_prank),
   # Compare % agreement between canaper and Biodiverse
   acacia_canape_comp = calc_agree_endem(
     df_1 = acacia_canape_cpr,
     df_2 = acacia_canape_biod
   ),
+  # Calculate time to run Biodiverse
+  tar_file_read(
+    acacia_biod_times_raw,
+    "_targets/user/data_raw/acacia_canape_biod_times.txt",
+    read_lines(!!.x)
+  ),
+  acacia_biod_elapsed_min = extract_biod_elapsed_min(acacia_biod_times_raw),
   # Render MS
   tar_quarto(
     ms_doc,

### 2022-09-29 _targets.R
Change n_iter for curveball

@@ -26,7 +26,7 @@ tar_plan(
     phy = acacia$phy,
     null_model = "curveball",
     n_reps = 999,
-    n_iterations = 100000,
+    n_iterations = 50000,
     workers = 2,
     hardware_info = get_mac_hardware_info(),
     req_spec = mac_laptop_specs
@@ -37,7 +37,7 @@ tar_plan(
     phy = acacia$phy,
     null_model = "curveball",
     n_reps = 999,
-    n_iterations = 100000,
+    n_iterations = 50000,
     workers = 3,
     hardware_info = get_mac_hardware_info(),
     req_spec = mac_laptop_specs
@@ -48,19 +48,10 @@ tar_plan(
     phy = acacia$phy,
     null_model = "curveball",
     n_reps = 999,
-    n_iterations = 100000,
+    n_iterations = 50000,
     workers = 1,
     hardware_info = get_mac_hardware_info(),
     req_spec = mac_laptop_specs
-  ),
-  # Run CANAPE using independent swap algorithm
-  acacia_canape_cpr_swap = run_canape(
-    comm = acacia$comm,
-    phy = acacia$phy,
-    null_model = "swap",
-    n_reps = 999,
-    n_iterations = 100000,
-    workers = 4
   ),
   # Load results from Biodiverse for Acacia
   # - p-rank results

### 2022-09-29 ms/ms.Qmd
Update MS, add SI

@@
 
-Running headline  
-*canaper*
+Running headline: *canaper*
 
@@
 
-3. Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 15 min to reproduce a canonical study).
+3. Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 10 min to reproduce a canonical study).
 
@@
 Community data is typically loaded using `read.csv()`, `readr::read_csv()`, or other functions that can import rectangular data.
+The `points2comm()` function of the `phyloregion` package may be used to convert raw occurrence data (e.g., latitude and longitude of species occurrences) to a community matrix [@Daru2020a].
 
@@
 
-The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with `canaper`, the test data from Phylocom [@Webb2008]:
+The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with `canaper`, the test data from Phylocom [@Webb2008].
 
@@
 It is generally impossible to know *a-priori* how many iterations are needed to completely randomize a given matrix; this number depends on properties of the matrix and the randomization algorithm.
-Generally, larger matrices with more skew (overabundance of zeros) will require more iterations (ADD CITATION).
+Generally, larger matrices with more skew (overabundance of zeros) will require more iterations [@Miklos2004].
 
@@
 
-The following code demonstrates usage of `cpr_rand_test()`, using the same example dataset as above:
+The following code demonstrates usage of `cpr_rand_test()`, using the same example dataset as above.
 
@@
 
-The following code demonstrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()` (note that for this small example, not all possible types of `endem_type` are produced):
+The following code demonstrates usage of `cpr_classify_endem()` with the output from `cpr_rand_test()` (note that for this small example, not all possible types of `endem_type` are produced).
 
@@
 
-The following code demonstrates parallel computing:
+The following code demonstrates parallel computing.
 
@@
 All `canaper` analyses were run with `canaper` v`r packageVersion("canaper")` in R v`r R.Version()$major`.`r R.Version()$minor`.
-Conducting CANAPE on Australian *Acacia* using `canaper` is simple, requiring only two commands:
+We used the `curveball` randomization algorithm with 50,000 iterations, since `cpr_iter_sim()` indicated this was sufficient to randomize the matrix (see "Major Functions").
+Once the randomization algorithm and its settings have been selected, CANAPE can be conducted with just two commands.
 
@@
   null_model = "curveball",
-  n_reps = 999, n_iterations = 100000,
+  n_reps = 999, n_iterations = 50000,
   tbl_out = TRUE
@@
 Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
-Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr):
+Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr).
 
@@
   coord_fixed() +
-  theme(legend.position = "bottom", legend.title = element_blank())
+  theme(
+    legend.position = "bottom",
+    legend.title = element_blank(),
+    panel.grid.minor = element_blank()
+  )
 
@@
   acacia_canape_biod,
-  acacia_canape_comp),
+  acacia_canape_comp,
+  acacia_biod_elapsed_min,
+  # Enforce dependency on acacia_canape_cpr_1 for calculating times
+  acacia_canape_cpr_1),
   store = here::here("_targets")
@@
 Palettes for plotting CANAPE and *p*-rank results can be selected with the `cpr_make_pal()` function.
+The same plot visualized with CVD-safe colors is available in the online supporting material (Fig. S1).
 
@@
 When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells (@tbl-cpr-biod).
-The total number of significant cells was nearly identical between the two (*n* = `r sig_tally$canaper$significant$n_sites` and `r sig_tally$biodiverse$significant$n_sites` sites, `canaper` and Biodiverse, respectively).
+The total number of significant cells was very similar (*n* = `r sig_tally$canaper$significant$n_sites` and `r sig_tally$biodiverse$significant$n_sites` sites, `canaper` and Biodiverse, respectively).
 The `rand_structured` and `curveball` null models both preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models [@Strona2018], which may explain the relative high agreement between results.
@@
 #| include: false
-
-# Load times
-tar_load(acacia_biod_elapsed_min)
-
-# Enforce dependency on acacia_canape_cpr_1
-tar_load(acacia_canape_cpr_1,
-  store = here::here("_targets"))
 
@@
 `adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.
-`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics but does not appear to be currently maintained and has been removed from CRAN.
+`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics.
+`EcoSimR` [@Gotelli2013a] provides randomization algorithms as well as functions for characterizing community matrices and null distributions.
+`PhyloMeasures` and `EcoSimR` have both been removed from CRAN and may not be currently maintained.
 
@@
 Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
+`canaper` has already been used at least three studies while in development [@Ellepola2022; @Lu2022; @Nitta2022a].
 We expect `canaper` will become a major tool in the toolkit of the emerging field of spatial phylogenetics alongside Biodiverse.
@@
 Source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
-Usage of `canaper` is documented at https://docs.ropensci.org/canaper/.
+Usage of `canaper` is documented at <https://docs.ropensci.org/canaper/>.
 The `acacia` dataset was originally published by @Mishler2014.
@@
 tibble(
-  Dataset = names(comms),
+  Dataset = names(comms) %>% paste0("`", ., "`"),
   `*n* sites` = map_dbl(comms, nrow),
   `*n* species (community matrix)` = map_dbl(comms, ncol),
-  `*n* species (tree)` = map_dbl(phylos, ape::Ntip)
+  `*n* species (phylogeny)` = map_dbl(phylos, ape::Ntip)
 ) %>%
@@
 
-ggsave(plot = fig, file = "ms/figure_1.pdf", width = 150, height = 150, units = "mm")
+ggsave(plot = fig, file = here::here("ms/figure_1.pdf"),
+  width = 150, height = 150, units = "mm")
 
@@
 
-ggsave(plot = fig, file = "ms/figure_2.pdf", width = 100, height = 200, units = "mm")
+ggsave(plot = fig, file = here::here("ms/figure_2.pdf"),
+  width = 100, height = 200, units = "mm")
 

### 2022-09-29 ms/si.Qmd
Update MS, add SI

---
title: "canaper: Categorical analysis of neo- and paleo-endemism in R"
subtitle: "Supporting information"
bibliography: references.yaml
csl: methods-in-ecology-and-evolution.csl
highlight-style: github
format:
  pdf:
    include-in-header:
      # keep figures from floating away
      text: |
        \usepackage{float}
        \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(targets)
library(tidyverse)
library(patchwork)
library(canaper)
```

```{r}
#| label: acacia-cpr
#| fig.height: 8
#| echo: false

acacia_canape <-
  tar_read(
    acacia_canape_cpr,
    store = here::here("_targets")
  ) |>
  separate(site, c("long", "lat"), sep = ":") |>
  mutate(across(c(long, lat), parse_number))

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  scale_fill_manual(values = cpr_endem_cols_4) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.background = element_rect(fill = "white", color = "white"),
    panel.grid.major = element_line(color = "grey60"),
    panel.grid.minor = element_blank()
  )

b <- ggplot(
  acacia_canape,
  aes(x = pe_alt_obs, y = pe_obs, color = endem_type)
) +
  geom_abline(slope = 1, color = "darkgrey") +
  geom_point() +
  scale_color_manual(values = cpr_endem_cols_4) +
  labs(
    x = "Phylogenetic endemism on comparison tree",
    y = "Phylogenetic endemism on actual tree"
  ) +
  theme_bw() +
  theme(legend.position = "none")

a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
```

Fig. S1: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors.
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type.

### 2022-09-29 R/functions.R

@@ -43,35 +43,6 @@ run_canape <- function(
   return(res)
 }
 
-#' Test number of optimal iterations needed for a given null model
-test_iter <- function(comm, null_model, n_iter, binary) {
-  if (isTRUE(binary)) {
-    # binary algorithms produce binary matrices, so convert input
-    # matrix to binary too
-    comm <- as.matrix(comm)
-    comm[comm > 0] <- 1
-  }
-  # Use a tibble to hold the data and run loops
-  tibble::tibble(
-    n_iter = n_iter,
-    # Make one random community for each value of `n_iter`
-    rand_comm = purrr::map(
-      n_iter,
-      ~ canaper::cpr_rand_comm(
-        comm,
-        null_model = null_model, n_iterations = .
-      )
-    ),
-    # Calculate the Pearson correlation (r) between each randomized
-    # matrix and the original matrix by converting
-    # the matrices to vectors
-    corr = map_dbl(
-      rand_comm,
-      ~ cor(c(comm), c(.)) # c() converts a matrix to vector
-    )
-  )
-}
-
 #' Classify endemism types for output from Biodiverse
 #'
 #' Description of Biodiverse output columns and corresponding canaper columns

### 2022-09-29 R/process_refs.R

@@ -5,10 +5,13 @@ library(rmdref)
 # Filter YAML references from Zotero library.
 # main_library.yaml has been exported from Zotero
 # like this: file -> "export library" -> "Better CSL YAML"
+# or created as a symlink like this:
+# cd _targets/user/data_raw/
+# ln -s ~/Dropbox/bibliography/main_library.yaml
 filter_refs_yaml(
   rmd_file = "ms/ms.Qmd",
   yaml_in = "_targets/user/data_raw/main_library.yaml",
-  yaml_out = "ms/references.yaml", silent = TRUE)
+  yaml_out = "ms/references.yaml", silent = FALSE)
 
 # Write out 'aux' file with citation keys to make collection in Zotero
 # so reference data are easier to work with.

### 2022-09-29 _targets.R

@@ -7,14 +7,6 @@ source("R/functions.R")
 tar_plan(
   # Load acacia dataset
   acacia = canaper::acacia,
-  # Test optimal number of iterations for making random communities
-  acacia_iter_test_res = test_iter(
-    comm = acacia$comm,
-    null_model = "curveball",
-    # test iterations from 10 to 10^7
-    n_iter = magrittr::raise_to_power(10, 1:7),
-    binary = TRUE
-  ),
   # Run CANAPE using canaper (2 cores) on mac laptop
   mac_laptop_specs = c(
     "Model Name: MacBook Pro",
@@ -84,8 +76,17 @@ tar_plan(
   acacia_biod_elapsed_min = extract_biod_elapsed_min(acacia_biod_times_raw),
   # Render MS
   tar_quarto(
-    ms_doc,
+    ms,
     "ms/ms.Qmd",
+    extra_files = c(
+      "ms/methods-in-ecology-and-evolution.csl",
+      "ms/references.yaml"
+    )
+  ),
+  # Render SI
+  tar_quarto(
+    si,
+    "ms/si.Qmd",
     extra_files = c(
       "ms/methods-in-ecology-and-evolution.csl",
       "ms/references.yaml"

### 2022-09-29 R/functions.R
Add set.seed()

@@ -11,10 +11,12 @@ get_mac_hardware_info <- function() {
 #' includes hardware specs when running on a Mac
 #' @param req_spec Character vector; strings to detect in `hardware_info`;
 #' if any are not detected, the function will fail
+#' @param seed The seed used for set.seed()
 run_canape <- function(
   comm, phy, null_model, n_reps = 500, n_iterations = 100000, workers = 6,
-  hardware_info = NULL, req_spec = NULL) {
+  hardware_info = NULL, req_spec = NULL, seed) {
   on.exit(plan(sequential))
+  set.seed(seed)
   # Check hardware requirements
   if (!is.null(hardware_info)) {
     hardware_check <- purrr::map_lgl(

### 2022-09-29 _targets.R

@@ -21,7 +21,8 @@ tar_plan(
     n_iterations = 50000,
     workers = 2,
     hardware_info = get_mac_hardware_info(),
-    req_spec = mac_laptop_specs
+    req_spec = mac_laptop_specs,
+    seed = 12345
   ),
   # Repeat, with three cores for comparison
   acacia_canape_cpr_3 = run_canape(
@@ -32,7 +33,8 @@ tar_plan(
     n_iterations = 50000,
     workers = 3,
     hardware_info = get_mac_hardware_info(),
-    req_spec = mac_laptop_specs
+    req_spec = mac_laptop_specs,
+    seed = 12345
   ),
   # Repeat, with one core for comparison
   acacia_canape_cpr_1 = run_canape(
@@ -43,7 +45,8 @@ tar_plan(
     n_iterations = 50000,
     workers = 1,
     hardware_info = get_mac_hardware_info(),
-    req_spec = mac_laptop_specs
+    req_spec = mac_laptop_specs,
+    seed = 12345
   ),
   # Load results from Biodiverse for Acacia
   # - p-rank results

### 2022-09-30 ms/ms.Qmd
Add set.seed()
Fix usage of seed in cpr_rand_test

@@
 # Load pre-calculated results from targets pipeline
-# - acacia_canape_cpr is Acacia CANAPE results calculated with canaper
+# - acacia_canape_cpr_1 is Acacia CANAPE results calculated with canaper
 # - rename it so we can use it in demo code below
-acacia_canape <- tar_read(acacia_canape_cpr,
+acacia_canape <- tar_read(acacia_canape_cpr_1,
   store = here::here("_targets"))
@@
 #| eval: false
+
+# Set a seed for reproducibility
+set.seed(12345)
 
@@
 tar_load(c(
-  acacia_canape_cpr,
   acacia_canape_biod,
@@
   acacia_biod_elapsed_min,
-  # Enforce dependency on acacia_canape_cpr_1 for calculating times
-  acacia_canape_cpr_1),
+  acacia_canape_cpr_1,
+  acacia_canape_cpr_2,
+  acacia_canape_cpr_3),
   store = here::here("_targets")
@@
 comp_table <- left_join(
-  count(acacia_canape_cpr, endem_type, name = "n_canaper"),
+  count(acacia_canape_cpr_1, endem_type, name = "n_canaper"),
   count(acacia_canape_biod, endem_type, name = "n_biodiverse"),
@@
 Approximate compute times were `r acacia_biod_elapsed_min` min for Biodiverse and `r times$acacia_canape_cpr_1$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
-However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr$minutes` min (further increases in number of cores did not significantly decrease time).
+However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr_2$minutes` min (further increases in number of cores did not significantly decrease time).
 This demonstrates that `canaper` can efficiently compute moderately sized datasets with a personal laptop computer.

### 2022-09-30 ms/si.Qmd
Fix usage of seed in cpr_rand_test

@@
   tar_read(
-    acacia_canape_cpr,
+    acacia_canape_cpr_1,
     store = here::here("_targets")

### 2022-09-30 R/functions.R

@@ -13,10 +13,9 @@ get_mac_hardware_info <- function() {
 #' if any are not detected, the function will fail
 #' @param seed The seed used for set.seed()
 run_canape <- function(
-  comm, phy, null_model, n_reps = 500, n_iterations = 100000, workers = 6,
+  comm, phy, null_model, n_reps = 500, n_iterations = 100000, workers = 1,
   hardware_info = NULL, req_spec = NULL, seed) {
-  on.exit(plan(sequential))
-  set.seed(seed)
+  on.exit(future::plan(future::sequential), add = TRUE)
   # Check hardware requirements
   if (!is.null(hardware_info)) {
     hardware_check <- purrr::map_lgl(
@@ -29,7 +28,12 @@ run_canape <- function(
     )
   }
   # Set up parallelization
-  plan(multisession, workers = workers)
+  if (workers > 1) {
+    future::plan(future::multisession, workers = workers)
+  } else {
+    future::plan(future::sequential)
+  }
+  set.seed(seed)
   # Run randomization test, returning output as tibble (not a dataframe)
   rand_res <- cpr_rand_test(
     comm = comm, phy = phy,
@@ -38,7 +42,7 @@ run_canape <- function(
     tbl_out = TRUE
   )
   # Unset parallelization
-  plan(sequential)
+  future::plan(future::sequential)
   # Classify endemism
   res <- cpr_classify_endem(rand_res) %>%
     mutate(site = str_remove_all(site, "'"))

### 2022-09-30 _targets.R

@@ -7,43 +7,44 @@ source("R/functions.R")
 tar_plan(
   # Load acacia dataset
   acacia = canaper::acacia,
-  # Run CANAPE using canaper (2 cores) on mac laptop
+  # Run CANAPE using canaper on mac laptop
+  # test up to three cores
   mac_laptop_specs = c(
     "Model Name: MacBook Pro",
     "Processor Name: Quad-Core Intel Core i7",
     "Total Number of Cores: 4",
     "Memory: 16 GB"),
-  acacia_canape_cpr = run_canape(
+  acacia_rep = 999,
+  acacia_iter = 50000,
+  acacia_canape_cpr_1 = run_canape(
     comm = acacia$comm,
     phy = acacia$phy,
     null_model = "curveball",
-    n_reps = 999,
-    n_iterations = 50000,
+    n_reps = acacia_rep,
+    n_iterations = acacia_iter,
+    workers = 1,
+    hardware_info = get_mac_hardware_info(),
+    req_spec = mac_laptop_specs,
+    seed = 12345
+  ),
+  acacia_canape_cpr_2 = run_canape(
+    comm = acacia$comm,
+    phy = acacia$phy,
+    null_model = "curveball",
+    n_reps = acacia_rep,
+    n_iterations = acacia_iter,
     workers = 2,
     hardware_info = get_mac_hardware_info(),
     req_spec = mac_laptop_specs,
     seed = 12345
   ),
-  # Repeat, with three cores for comparison
   acacia_canape_cpr_3 = run_canape(
     comm = acacia$comm,
     phy = acacia$phy,
     null_model = "curveball",
-    n_reps = 999,
-    n_iterations = 50000,
+    n_reps = acacia_rep,
+    n_iterations = acacia_iter,
     workers = 3,
-    hardware_info = get_mac_hardware_info(),
-    req_spec = mac_laptop_specs,
-    seed = 12345
-  ),
-  # Repeat, with one core for comparison
-  acacia_canape_cpr_1 = run_canape(
-    comm = acacia$comm,
-    phy = acacia$phy,
-    null_model = "curveball",
-    n_reps = 999,
-    n_iterations = 50000,
-    workers = 1,
     hardware_info = get_mac_hardware_info(),
     req_spec = mac_laptop_specs,
     seed = 12345
@@ -67,7 +68,7 @@ tar_plan(
     biod_res_prank = acacia_biod_res_prank),
[…]

### 2022-09-30 R/workflow_hist_plot.R
Add script to make histogram for workflow fig

@@ -1,0 +1,26 @@
+# Make a histogram with shaded areas
+
+library(tidyverse)
+library(ggpattern)
+
+# see https://r-graphics.org/recipe-miscgraph-function-shade
+dnorm_limit <- function(x) {
+    y <- dnorm(x)
+    y[x > -2 & x < 2] <- NA
+    return(y)
+}
+
+# ggplot() with dummy data
+p <- ggplot(data.frame(x = c(-3, 3)), aes(x = x))
+
+p +
+  stat_function(fun = dnorm_limit, geom = "area", fill = "grey", alpha = 0.8) +
+  stat_function(fun = dnorm, lwd = 1.2) +
+  theme_classic(base_size = 40) +
+  theme(
+    axis.text = element_blank(),
+    axis.ticks = element_blank(),
+    axis.title = element_blank()
+    )
+
+ggsave("working/hist.png", dpi = 600)

### 2022-09-30 ms/ms.Qmd
Update MS

@@
 format:
-  docx: default
+  docx:
+    reference-doc: custom-reference.docx
   pdf:
@@
 Parallel computing can be enabled with a single line of code.
-`canaper` has undergone thorough code review through ROpenSci [@Boettiger2015] and is verified against a large number of unit tests (>95% coverage).
+`canaper` has passed code review meeting the "silver" standards for statistical software at ROpenSci [@Boettiger2015], and is verified against a large number of unit tests (>99% coverage).
 All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.
@@
 The entire CANAPE workflow can be run with two functions, `cpr_rand_test()` and `cpr_classify_endem()`.
-However, internally this entails several steps that the user should be aware of as follows.
+However, internally this entails several steps that the user should be aware of as follows (@fig-workflow).
 
@@
 Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
-Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr).
+Here, we demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr).
 
@@
 However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
-Alternate color palettes (e.g., `cpr_endem_cols`) are also vailable that include CVD-safe colors [@Okabe2002].
+Alternate color palettes (e.g., `cpr_endem_cols`) are also available in `canaper` that include CVD-safe colors [@Okabe2002].
 Palettes for plotting CANAPE and *p*-rank results can be selected with the `cpr_make_pal()` function.
@@
 `EcoSimR` [@Gotelli2013a] provides randomization algorithms as well as functions for characterizing community matrices and null distributions.
-`PhyloMeasures` and `EcoSimR` have both been removed from CRAN and may not be currently maintained.
+As of writing, `PhyloMeasures` and `EcoSimR` have both been removed from CRAN and may not be currently maintained.
 
@@
 All data and code are available in the `canaper` R package, which is distributed by the Comprehensive R Archive Network (CRAN).
-Source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
+`canaper` source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
 Usage of `canaper` is documented at <https://docs.ropensci.org/canaper/>.
+Code and data used to run analyses and generate this manuscript available at <https://github.com/joelnitta/canaper_ms>.
 The `acacia` dataset was originally published by @Mishler2014.
@@
 
+**@fig-workflow.** `canaper` workflow and functions to conduct categorical analysis of neo- and paleo-endemism (CANAPE).
+Input data are a community matrix and phylogeny.
+First, the user selects a randomization algorithm and its settings, with optional assistance from the `cpr_iter_sim()` function.
+Next, a set of random communities are generated, various metrics (e.g., phylogenetic endemism, relative phylogenetic endemism) are calculated for the input community and each random community, and the rank (*p*-value) of each observed metric is compared to those of the random communities by the `cpr_rand_test()` function.
+Finally, endemism types are categorized on the basis of the ranked metrics with the `cpr_classify_endem()` function.
+For details on metrics and classification, see "Analysis Workflow".
+For details on functions, see "Major Functions".
+
 **@fig-iter-sim.** Testing the number of iterations needed to randomize the Australian *Acacia* dataset.
@@
 (a) Map of Australia showing grid-cells (communities) colored by endemism type.
+Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system [@Butler2007]; units in m.
 (b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.
@@
 
+{{< pagebreak >}}
+
+![`canaper` workflow and functions to conduct categorical analysis of neo- and paleo-endemism (CANAPE)](workflow.png){#fig-workflow}
+
 ```{r}

### 2022-09-30 _targets.R

@@ -84,7 +84,8 @@ tar_plan(
     "ms/ms.Qmd",
     extra_files = c(
       "ms/methods-in-ecology-and-evolution.csl",
-      "ms/references.yaml"
+      "ms/references.yaml",
+      "ms/custom-reference.docx"
     )
   ),
   # Render SI

### 2022-09-30 R/workflow_hist_plot.R
Don't need ggpattern

@@ -1,7 +1,5 @@
 # Make a histogram with shaded areas
-
 library(tidyverse)
-library(ggpattern)
 
 # see https://r-graphics.org/recipe-miscgraph-function-shade
 dnorm_limit <- function(x) {

### 2022-09-30 ms/ms.Qmd
Update MS

@@
 
-Joel H. Nitta^1,\*^, Shawn W. Laffan^2^, Brent D. Mishler^3^, and Wataru Iwasaki^1,4,5,6,7^
+Joel H. Nitta^1,\*^, Shawn W. Laffan^2^, Brent D. Mishler^3^, and Wataru Iwasaki^1,4,5,6,7,8^
 
@@
 
-^5^Atmosphere and Ocean Research Institute, The University of Tokyo, Chiba, Japan
+^5^Department of Biological Sciences, Graduate School of Science, The University of Tokyo, Tokyo, Japan
 
-^6^Institute for Quantitative Biosciences, The University of Tokyo, Tokyo, Japan
+^6^Atmosphere and Ocean Research Institute, The University of Tokyo, Chiba, Japan
 
-^7^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan
+^7^Institute for Quantitative Biosciences, The University of Tokyo, Tokyo, Japan
+
+^8^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan
 
@@
 All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.
+
+# Installation
+
+The stable version of `canaper` can be installed from CRAN, and the latest development version from r-universe (<https://r-universe.dev/>).
+
+```{r}
+#| label: installation
+#| eval: false
+
+# Stable version
+install.packages("canaper")
+
+# Development version
+install.packages("canaper", repos = "https://ropensci.r-universe.dev")
+```
 
@@
 
-data(phylocom)
+# Set a seed for reproducibility
+set.seed(12345)
 
 # The original data includes abundance
+data(phylocom)
 phylocom$comm
@@
 #| include: false
+set.seed(12345)
+
 data(acacia)
@@
 #| label: cpr-iter-sim-demo
+set.seed(12345)
+
 data(acacia)
@@
 #| label: cpr-rand-test-demo
+set.seed(12345)
+
 rand_test_results <- cpr_rand_test(
@@
 library(future)
+set.seed(12345)
 
@@
   geom_tile() +
-  # use same colors as Misher et al. 2014
-  # (for CVD-safe colors see cpr_endem_cols or cpr_endem_cols_2)
+  # Use same colors as Misher et al. 2014
+  # (for CVD-safe colors see cpr_endem_cols)
   scale_fill_manual(values = mishler_endem_cols) +
@@
 
-# Tally number of significant cells in canper vs. Biodiverse
+# Tally number of significant cells in canaper vs. Biodiverse
 sig_tally <-
@@
 
-This research supported in part by Japan Society for the Promotion of Science (Kakenhi) Grant numbers 16H06279 and 22H04925.
+This research supported in part by Japan Society for the Promotion of Science (Kakenhi) Grant numbers 16H06279, 22H04925, and 22K15171.
 Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
@@
 
-Joel H. Nitta conceived the ideas and designed methodology; Joel H. Nitta and Shawn W. Laffan analyzed the data; Joel H. Nitta led the writing of the manuscript.
+Joel H.
+Nitta conceived the ideas and designed methodology; Joel H.
+Nitta and Shawn W.
+Laffan analyzed the data; Joel H.
+Nitta led the writing of the manuscript; Brent D.
+Mishler edited the manuscript; Wataru Iwasaki provided resources and supervised the study.
 All authors contributed critically to the drafts and gave final approval for publication.

### 2022-10-05 ms/ms.Qmd
Update MS

@@
 
-^2^Department of Integrative Biology, University of California Berkeley, Berkeley, USA
+^2^School of Biological, Earth and Environmental Sciences, The University of New South Wales, Kensington, Australia
 
-^3^School of Biological, Earth and Environmental Sciences, The University of New South Wales, Kensington, Australia
+^3^Department of Integrative Biology, University of California Berkeley, Berkeley, USA
 
@@
 CANAPE (categorical analysis of neo- and paleo-endemism) is one such method that can provide insight into the evolutionary processes generating biodiversity.
-The only currently available software implementing CANAPE is Biodiverse, which is written in perl and can be used either through a graphical user interface (GUI) or user-developed scripts.
+The only currently available software implementing CANAPE is Biodiverse, which is written in Perl and can be used either through a graphical user interface (GUI) or user-developed scripts.
 However, many researchers, particularly in the fields of ecology and evolutionary biology, use the R programming language to conduct their analyses.
@@
 Quantifying biodiversity is a major goal of ecology.
-Until recently, quantification of biology was done on a purely taxonomic basis: counting the number of species in an area, for example [e.g., @Diamond1975].
+Quantification of biodiversity is often done on a purely taxonomic basis: counting the number of species in an area [e.g., @Diamond1975].
 However, as all taxa are related to some degree by descent from a common ancestor, a thorough understanding of biodiversity is only possible by considering their evolutionary relationships.
@@
 One recently developed extension of PE is categorical analysis of neo-and paleo-endemism [CANAPE\; @Mishler2014].
-CANAPE uses phylogenetic methods to give insight into evolutionary processes that produce endemism.
-In theory, centers of endemism (areas with many range-limited taxa; or in the phylogenetic sense, concentrations of range-limited branches of the phylogeny) may arise via multiple processes.
-For example, previously widespread lineages may undergo extinction in all but a portion of their range (paleo-endemism), or recently diverged lineages may only occur in a small area (neo-endemism). 
+CANAPE uses phylogenetic methods to give insight into evolutionary processes that produce centers of endemism (areas with many narrow-ranged taxa; or in the phylogenetic sense, concentrations of narrow-ranged branches of the phylogeny).
+In theory, centers of endemism may arise via multiple processes.
+For example, previously widespread lineages may undergo extinction in all but a portion of their range, leading to paleo-endemism.
+Alternately, recently diverged lineages may only occur in a small area and lead to neo-endemism.
 It is also possible that a given area is home to a high concentration of both paleo- and neo-endemic lineages (mixed endemism).
-CANAPE involves analyzing observed patterns of PE in comparison with a null model to infer first whether an area is significantly high in PE, then classifying those areas into centers of paleo-endemism, neo-endemism, or mixed endemism.
+CANAPE involves analyzing observed patterns of PE in comparison with a null model to first infer whether an area is significantly high in PE, and then to classify those areas into centers of paleo-endemism, neo-endemism, or mixed endemism.
 CANAPE is widely used (\> 50 publications in Google Scholar query for papers that cite @Mishler2014 and mention "CANAPE"), and is a central component of the field of spatial phylogenetics.
@@
 Despite the popularity of CANAPE, it has so far only been implemented in one software package, Biodiverse [@Laffan2010].
-Biodiverse is written in perl and comprises an analytical engine and a graphical user interface (GUI).
+Biodiverse is written in Perl and comprises an analytical engine and a graphical user interface (GUI).
 While Biodiverse is convenient for non-coders because of its GUI, many ecologists and evolutionary biologists use R for their analyses [@Lai2019].
 Until now, an R user who wanted to conduct CANAPE analysis as part of a broader R workflow needed to first clean raw data, export it to Biodiverse, conduct PD and PE analyses in Biodiverse, then import the results back into R for further analysis and visualization.
-Also, while Biodiverse conducted the metrics needed to categorize endemism type, it did not actually perform this categorization (although this feature is planned to be added to a future version of Biodiverse).
+Also, while Biodiverse calculated the metrics needed to categorize endemism type, it did not actually perform this categorization (although this feature will be included in a future version of Biodiverse).
 Furthermore, the Biodiverse GUI alone does not support parallel processing, which is needed for large datasets.
-Parallel processing and automation of Biodiverse analyses is only possible using perl scripts.
-A set of R scripts is available to call Biodiverse perl scripts from R (<https://github.com/NunzioKnerr/biodiverse_pipeline>), but it is not an R package and does not conduct CANAPE within R.
+Parallel processing and automation of Biodiverse analyses is only currently possible using Perl scripts.
+A set of R scripts is available to call Biodiverse Perl scripts from R (<https://github.com/NunzioKnerr/biodiverse_pipeline>), but it is not an R package and does not conduct CANAPE within R.
 
 Here, we present a new R package that implements CANAPE completely in R: `canaper`.
-We strived to make `canaper` simple to use and efficient.
+We strove to make `canaper` simple to use and efficient.
 Parallel computing can be enabled with a single line of code.
@@
 
-Community data is provided as a `data.frame` or matrix, with species as columns and communities (also referred to as "sites" or "grid-cells") as rows.
+Community data is provided as a data frame or matrix, with species as columns and communities (also referred to as "sites" or "grid-cells") as rows.
 In this case, the data must include both row names and column names.
@@
 To minimize memory requirements, the individual random communities are not stored in memory.
-Since 100s to 1000s or more random replicates may be required, the memory usage could be quite high for large matrices if each replicate was stored.
+Since hundreds to thousands or more random replicates may be required, the memory usage could be quite high for large matrices if each replicate was stored.
 Rather, as each random community is generated, the summary statistics are computed immediately and stored.
-This way, for any input matrix of size $N \times M$ and $n$ random replicates, where $N$ is number of rows (sites) and $M$ is number columns (species), $n$ summary matrices (dataframes) each with $N$ rows and one column per summary statistic are stored, rather than $n$ matrices of size $N \times M$ (the same approach is used in Biodiverse).
+This way, for any input matrix of size $N \times M$ and $n$ random replicates, where $N$ is number of rows (sites) and $M$ is number columns (species), $n$ summary matrices (data frames) each with $N$ rows and one column per summary statistic are stored, rather than $n$ matrices of size $N \times M$ (the same approach is used in Biodiverse).
 
@@
 The `cpr_rand_comm()` function generates a single random community. 
-The first argument, `comm` is a community dataframe (or matrix). 
-The second, `null_model` is the name of one of the predefined randomization algorithms available in `vegan`.
+The first argument, `comm`, is a community data frame (or matrix). 
+The second, `null_model`, is the name of one of the predefined randomization to use.
 The remainder of the arguments are particular to specific types of randomization algorithm.
@@
 Binary algorithms are designed for binary (i.e., presence-absence) data, and quantitative algorithms are designed for quantitative (i.e., abundance) data.
-Either type of algorithm will accept either type of data, but binary algorithms will convert abundance data to binary and return a binary matrix (dataframe).
+Either type of algorithm will accept either type of data, but binary algorithms will convert abundance data to binary and return a binary matrix (data frame).
 
@@
 The following code demonstrates usage of `cpr_rand_test()` with the *Acacia* dataset [@Mishler2014] that comes with `canaper`.
-The *Acacia* dataset is large (@tbl-datasets) and highly skewed (`r acacia_non_empty` zeros).
+The *Acacia* dataset is relatively large (@tbl-datasets) and highly skewed (`r acacia_non_empty` zeros).
 
@@
 
-The output is a dataframe with communities as rows and summary statistics in columns.
+The output is a data frame with communities as rows and summary statistics in columns.
 A large number of summary statistics, including all of those needed to calculate CANAPE, are produced.
@@
 The `cpr_classify_endem()` function classifies endemism types for the output of `cpr_rand_test()` as described above in "Analysis workflow".
-The input is a dataframe including the following columns calculated by `cpr_rand_test()`: `pe_obs_p_upper` (upper *p*-value comparing observed PE to random values), `pe_alt_obs_p_upper` (upper *p*-value comparing observed PE on alternate tree to random values), `rpe_obs_p_upper` (upper *p*-value for RPE), and `rpe_obs_p_lower` (lower *p*-value for RPE).
-The output is the same dataframe, with the column `endem_type` appended.
+The input is a data frame including the following columns calculated by `cpr_rand_test()`: `pe_obs_p_upper` (upper *p*-value comparing observed PE to random values), `pe_alt_obs_p_upper` (upper *p*-value comparing observed PE on alternate tree to random values), `rpe_obs_p_upper` (upper *p*-value for RPE), and `rpe_obs_p_lower` (lower *p*-value for RPE).
+The output is the same data frame, with the column `endem_type` appended.
 Values of `endem_type` include `paleo` (paleoendemic), `neo` (neoendemic), `not significant`, `mixed` (mixed endemism), and `super` (super-endemic).
@@
 In addition to `cpr_classify_endem()`, `canaper` includes the `cpr_classify_signif()` function for assessing significance level of PD, RPD, PE, and RPE.
-Both `cpr_classify_endem()` and `cpr_classify_signif()` take a dataframe as their first argument and return a dataframe as output, so they are "pipe friendly", i.e., can be chained together using the pipe (`%>%` or `|>`).
+Both `cpr_classify_endem()` and `cpr_classify_signif()` take a data frame as their first argument and return a data frame as output, so they are "pipe friendly", i.e., can be chained together using pipe operators (`%>%` or `|>`).
 
@@
 `canaper` comes with three datasets for testing and demonstration (@tbl-datasets).
-Each is a list with two named elements, `comm` (community matrix, dataframe) and `phy` (phylogenetic tree, list of class `"phylo"`).
+Each is a list with two named elements, `comm` (community matrix, data frame) and `phy` (phylogenetic tree, list of class `"phylo"`).
 `acacia` includes data of Australian *Acacia* analyzed by @Mishler2014.
@@
 Approximate compute times were `r acacia_biod_elapsed_min` min for Biodiverse and `r times$acacia_canape_cpr_1$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
-However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr_2$minutes` min (further increases in number of cores did not significantly decrease time).
+However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr_2$minutes` min.
 This demonstrates that `canaper` can efficiently compute moderately sized datasets with a personal laptop computer.
@@
 `EcoSimR` [@Gotelli2013a] provides randomization algorithms as well as functions for characterizing community matrices and null distributions.
-As of writing, `PhyloMeasures` and `EcoSimR` have both been removed from CRAN and may not be currently maintained.
+As of writing, `PhyloMeasures` and `EcoSimR` had been removed from CRAN and may not be under active development.
 
@@
 
-All data and code are available in the `canaper` R package, which is distributed by the Comprehensive R Archive Network (CRAN).
-`canaper` source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
+The `canaper` R package is distributed by the Comprehensive R Archive Network (CRAN), with source code available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
 Usage of `canaper` is documented at <https://docs.ropensci.org/canaper/>.
-Code and data used to run analyses and generate this manuscript available at <https://github.com/joelnitta/canaper_ms>.
+Code and data used to conduct analyses and generate this manuscript are available at <https://github.com/joelnitta/canaper_ms>, and can be run using the Docker image `joelnitta/canaper_ms` available at <https://hub.docker.com/r/joelnitta/canaper_ms>.
 The `acacia` dataset was originally published by @Mishler2014.
@@
 - Community matrix: A 2-dimensional dataset with communties (grid-cells) on one axis and species on the other. `canaper` expects communities as rows and species as columns.
-- Phylogenetic diversity (PD): The total branch length (including the root) of a phylogenetic tree that includes all species occurring in a given area (community or grid-cell) [@Faith1992].
-- Phylogenetic endemism (PE): PD inversely weighted by the distribution range (i.e., number of grid-cells) of a branch [@Rosauer2009].
+- Phylogenetic diversity (PD): The total branch length (including the root) connecting the species occurring in a given area (community or grid-cell), measured on an overall tree that includes all the species in the study [@Faith1992].
+- Phylogenetic endemism (PE): Range weighted PD where the weight for each branch is the fraction of its overall range represented in the sample area (typically a grid-cell) [@Rosauer2009].
 - Relative phylogenetic diversity (RPD): The ratio of PD measured on the original tree vs. PD measured on a comparison tree with all branch lengths transformed to equal length [@Mishler2014].
 - Relative phylogenetic endemism (RPE): The ratio of PE measured on the original tree vs. PE measured on a comparison tree with all branch lengths transformed to equal length [@Mishler2014].
-- Neoendemic: An area with a high concentration of short, range-restricted branchlengths; may be due to processes such as recent speciation (radiation).
-- Paleoendemic: An area with a high concentration of long, range-restricted branchlengths; may be due to processes such as extinction or colonization by distantly related species.
-- Mixed endemic: An area with a mixture of short and long, range-restricted branchlengths; may be due to multiple processes.
+- Neoendemic: An area with a high concentration of short, range-restricted branches; may be due to processes such as recent speciation (radiation).
+- Paleoendemic: An area with a high concentration of long, range-restricted branches; may be due to processes such as extinction or colonization by distantly relatives from outside the study area.
+- Mixed endemic: An area with a mixture of short and long, range-restricted branches; may be due to multiple processes.
 - Super endemic: An area with mixed endemism that is statistically highly significant.
 
-**Box 2**: Columns in dataframe produced by `cpr_rand_test()`.
+**Box 2**: Columns in data frame produced by `cpr_rand_test()`.
 

### 2022-10-05 install_latex.R
Add dockerfile

@@ -1,0 +1,33 @@
+# Install latex packages using tinytex
+
+# This would happen automatically anyways when rendering the Rmd to pdf, 
+# but requires downloading packages and may not work during updates of Tex Live. 
+# Better to install to the docker image once and keep them there.
+
+tinytex::tlmgr_update()
+
+latex_packages <- c(
+  "koma-script",
+  "amsmath",
+  "latex-amsmath-dev",
+  "iftex",
+  "unicode-math",
+  "xcolor",
+  "fancyvrb",
+  "booktabs",
+  "etoolbox",
+  "mdwtools",
+  "float",
+  "caption",
+  "tcolorbox",
+  "pgf",
+  "environ",
+  "pdfcol",
+  "ltxcmds",
+  "infwarerr",
+  "hyperref",
+  "kvoptions",
+  "bookmark"
+)
+
+tinytex::tlmgr_install(latex_packages)

### 2022-10-05 ms/ms.Qmd
Update targets plan
get mac hardware info outside of docker

@@
   acacia_canape_cpr_1,
-  acacia_canape_cpr_2,
-  acacia_canape_cpr_3),
+  acacia_canape_cpr_2),
   store = here::here("_targets")

### 2022-10-05 R/get_mac_hardware_info.R

@@ -1,0 +1,8 @@
+# Get hardware info on a mac
+# Run this outside of the docker image before running _targets.R
+source(here::here("R/packages.R"))
+source(here::here("R/functions.R"))
+
+saveRDS(
+  get_mac_hardware_info(),
+  here::here("_targets/user/data_raw/my_hardware_info.RDS"))

### 2022-10-05 _targets.R

@@ -8,7 +8,9 @@ tar_plan(
   # Load acacia dataset
   acacia = canaper::acacia,
   # Run CANAPE using canaper on mac laptop
-  # test up to three cores
+  # - must run R/get_mac_hardware_info.R first outside of Docker
+  #   OR, set `hardware_info` and `req_spec` = NULL
+  #   to skip hardware requirements
   mac_laptop_specs = c(
     "Model Name: MacBook Pro",
     "Processor Name: Quad-Core Intel Core i7",
@@ -23,7 +25,7 @@ tar_plan(
     n_reps = acacia_rep,
     n_iterations = acacia_iter,
     workers = 1,
-    hardware_info = get_mac_hardware_info(),
+    hardware_info = readRDS("_targets/user/data_raw/my_hardware_info.RDS"),
     req_spec = mac_laptop_specs,
     seed = 12345
   ),
@@ -34,18 +36,7 @@ tar_plan(
     n_reps = acacia_rep,
     n_iterations = acacia_iter,
     workers = 2,
-    hardware_info = get_mac_hardware_info(),
-    req_spec = mac_laptop_specs,
-    seed = 12345
-  ),
-  acacia_canape_cpr_3 = run_canape(
-    comm = acacia$comm,
-    phy = acacia$phy,
-    null_model = "curveball",
-    n_reps = acacia_rep,
-    n_iterations = acacia_iter,
-    workers = 3,
-    hardware_info = get_mac_hardware_info(),
+    hardware_info = readRDS("_targets/user/data_raw/my_hardware_info.RDS"),
     req_spec = mac_laptop_specs,
     seed = 12345
   ),

### 2022-10-05 R/get_mac_hardware_info.R
Simplify code to get hardware info

@@ -1,8 +1,6 @@
 # Get hardware info on a mac
 # Run this outside of the docker image before running _targets.R
-source(here::here("R/packages.R"))
-source(here::here("R/functions.R"))
 
 saveRDS(
-  get_mac_hardware_info(),
-  here::here("_targets/user/data_raw/my_hardware_info.RDS"))
+  system("system_profiler SPHardwareDataType", intern = TRUE),
+  "_targets/user/data_raw/my_hardware_info.RDS")

### 2022-10-06 ms/ms.Qmd
Update MS

@@
   tbl-title: Table
+  fig-prefix: Fig.
+  tbl-prefix: Table
 format:
@@
 Parallel computing can be enabled with a single line of code.
-`canaper` has passed code review meeting the "silver" standards for statistical software at ROpenSci [@Boettiger2015], and is verified against a large number of unit tests (>99% coverage).
+`canaper` has passed code review meeting the "silver" standards for statistical software at rOpenSci [@Boettiger2015], and is verified against a large number of unit tests (>99% coverage).
 All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.
@@
 
-Once a randomization algorithm has been selected, random communities are generated for a number of replicates set by the user, and a set of summary statistics are computed (Box 2).
+Once a randomization algorithm has been selected, random communities are generated for a number of replicates set by the user, and a set of summary statistics are computed (Appendix S1).
 Summary statistics include the mean and standard deviation of PD, RPD, PE, and RPE of the random communities and comparisons of observed values to the random communities including standard effect size and rank, which is used to calculate *p*-values.
-
-To minimize memory requirements, the individual random communities are not stored in memory.
-Since hundreds to thousands or more random replicates may be required, the memory usage could be quite high for large matrices if each replicate was stored.
-Rather, as each random community is generated, the summary statistics are computed immediately and stored.
-This way, for any input matrix of size $N \times M$ and $n$ random replicates, where $N$ is number of rows (sites) and $M$ is number columns (species), $n$ summary matrices (data frames) each with $N$ rows and one column per summary statistic are stored, rather than $n$ matrices of size $N \times M$ (the same approach is used in Biodiverse).
 
@@
 The final step in CANAPE is to categorize endemism as described in @Mishler2014.
-Briefly, this is done by comparing significance values of summary statistics calculated in the previous step (Box 3).
+Briefly, this is done by comparing significance values of summary statistics calculated in the previous step (Appendix S2).
 To be considered significantly endemic, a given grid-cell must first have significantly high `pe_obs` or `pe_alt` or both (one-tailed test).
@@
 
-`cpr_iter_sim()` helps the user determine the minimum number of iterations needed by conducting successive exchanges and recording the similarity between the original matrix and the randomized matrix at each iteration.
+`cpr_iter_sim()` conducts successive exchanges and records the similarity between the original matrix and the randomized matrix at each iteration.
 The similarity values should initially decrease until an approximate minimum is reached; further iterations will only result in noise around this minimum.
@@
 A large number of summary statistics, including all of those needed to calculate CANAPE, are produced.
-For a full explanation of all output columns, see Box 2 or run `?cpr_rand_test()`.
+For a full explanation of all output columns, see Appendix S1 or run `?cpr_rand_test()`.
 
@@
 There will not be a noticeable decrease in computing time since this dataset is so small.
-However, parallelization can greatly decrease computing time for large datasets; more details are available in the "Parallel computing" vignette (<https://joelnitta.github.io/canaper/articles/parallel.html>).
+However, parallelization can greatly decrease computing time for large datasets; more details are available in the "Parallel computing" vignette (<https://docs.ropensci.org/canaper/articles/parallel.html>).
 
@@
 `acacia` includes data of Australian *Acacia* analyzed by @Mishler2014.
-`phylocom` and `biod_example` are both fictitious datasets compiled for testing software, and were originally published in Phylocom [@Webb2008] and Biodiverse [@Laffan2010], respectively.
+`phylocom` and `biod_example` are fictitious datasets compiled for testing software, and were originally published in Phylocom [@Webb2008] and Biodiverse [@Laffan2010], respectively.
 
@@
 `vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
-`vegan` includes by far the greatest number of algorithms for generating random communities (> 30) and is used by `canaper`.
+`vegan` includes by far the greatest number of algorithms for generating random communities and is used by `canaper`.
 `adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.
@@
 Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
-Luis Osorio and Klaus Schliep reviewed the code as part of submission to ROpenSci.
+Luis Osorio and Klaus Schliep reviewed the code as part of submission to rOpenSci.
 
@@
 
-**Box 2**: Columns in data frame produced by `cpr_rand_test()`.
-
-`cpr_rand_test()` calculates six basic metrics for each grid-cell (Box 1):
-
-- `pd`: phylogenetic diversity (PD).
-- `pd_alt`: alternative PD (PD measured on comparison tree with all branch lengths transformed to equal length).
-- `pe`: phylogenetic endemism (PE).
-- `pe_alt`: alternative PE (PE measured on comparison tree).
-- `rpd`: relative PD (ratio of `pd` to `pd_alt`).
-- `rpe`: relative PE (ratio of `pe` to `pe_alt`).
-
-For each of the six basic metrics, nine columns (54 total) are output by `cpr_rand_test()`:
-
-- `*_obs`: Observed value.
-- `*_obs_c_lower`: Count of times observed value was lower than random values.
-- `*_obs_c_upper`: Count of times observed value was higher than random values.
-- `*_obs_p_lower`: Percentage of times observed value was lower than random values.
-- `*_obs_p_upper`: Percentage of times observed value was higher than random values.
-- `*_obs_q`: Count of the non-NA random values used for comparison.
-- `*_obs_z`: Standard effect size (*z*-score).
-- `*_rand_mean`: Mean of the random values.
-- `*_rand_sd`: Standard deviation of the random values.
-
-So for `pd` the output columns would include `pd_obs`, `pd_obs_c_lower`, etc.
-The *p*-values (`*_obs_p_upper`, etc.) are then used to categorize endemism type according to CANAPE (Box 3).
-
-**Box 3**: Criteria for categorical analysis of neo- and paleo-endemism (CANAPE).
-
-Significance assessed at $\alpha = 0.05$ (significant) or $\alpha = 0.01$ (highly significant).
-Metrics described in Box 2.
-Not all studies include super-endemism.
-
-- If either `pe_obs_p_upper` or `pe_alt_obs_p_upper` are significantly high (one-sided test), then look for paleo- or neo-endemism.
-  - If `rpe_obs_p_upper` is significantly high (two-sided test), then grid-cell is paleo-endemic.
-  - Else if `rpe_obs_p_lower` is significantly low (two-sided test), then grid-cell is neo-endemic.
-  - Else there is mixed age endemism, in which case:
-    - If both `pe_obs_p_upper` and `pe_alt_obs_p_upper` are highly significant (*p* < 0.01), then grid-cell is super-endemic.
-    - Else grid-cell is mixed-endemic.
-- Else if neither `pe_obs_p_upper` nor `pe_alt_obs_p_upper` are significantly high, grid-cell is non-endemic.
-
 {{< pagebreak >}}
@@
 
-ggsave(plot = fig, file = here::here("ms/figure_1.pdf"),
+ggsave(plot = fig, file = here::here("ms/figure_2.pdf"),
   width = 150, height = 150, units = "mm")
@@
 
-ggsave(plot = fig, file = here::here("ms/figure_2.pdf"),
+ggsave(plot = fig, file = here::here("ms/figure_3.pdf"),
   width = 100, height = 200, units = "mm")

### 2022-10-06 ms/si.Qmd
Update MS

@@
 subtitle: "Supporting information"
+author:
+  - Joel H. Nitta
+  - Shawn W. Laffan
+  - Brent D. Mishler
+  - Wataru Iwasaki
 bibliography: references.yaml
@@
 
-Fig. S1: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors.
+**Fig. S1**: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors.
 (a) Map of Australia showing grid-cells (communities) colored by endemism type.
+Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system; units in m.
 (b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.
 a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type.
+
+{{< pagebreak >}}
+
+**Appendix S1**: Columns in data frame produced by `cpr_rand_test()`.
+
+`cpr_rand_test()` calculates six basic metrics for each grid-cell:
+
+- `pd`: phylogenetic diversity (PD).
+- `pd_alt`: alternative PD (PD measured on comparison tree with all branch lengths transformed to equal length).
+- `pe`: phylogenetic endemism (PE).
+- `pe_alt`: alternative PE (PE measured on comparison tree).
+- `rpd`: relative PD (ratio of `pd` to `pd_alt`).
+- `rpe`: relative PE (ratio of `pe` to `pe_alt`).
+
+For each of the six basic metrics, nine columns (54 total) are output by `cpr_rand_test()`:
+
+- `*_obs`: Observed value.
+- `*_obs_c_lower`: Count of times observed value was lower than random values.
+- `*_obs_c_upper`: Count of times observed value was higher than random values.
+- `*_obs_p_lower`: Percentage of times observed value was lower than random values.
+- `*_obs_p_upper`: Percentage of times observed value was higher than random values.
+- `*_obs_q`: Count of the non-NA random values used for comparison.
+- `*_obs_z`: Standard effect size (*z*-score).
+- `*_rand_mean`: Mean of the random values.
+- `*_rand_sd`: Standard deviation of the random values.
+
+So for `pd` the output columns would include `pd_obs`, `pd_obs_c_lower`, etc.
+The *p*-values (`*_obs_p_upper`, etc.) are then used to categorize endemism type according to CANAPE (Appendix S2).
+
+To minimize memory requirements, the individual random communities are not stored in memory when calculating metrics.
+Since hundreds to thousands or more random replicates may be required, the memory usage could be quite high for large matrices if each replicate was stored.
+Rather, as each random community is generated, the summary statistics are computed immediately and stored (the same approach is used in Biodiverse).
+
+{{< pagebreak >}}
+
+**Appendix S2**: Criteria for categorical analysis of neo- and paleo-endemism (CANAPE).
+
+Significance assessed at $\alpha = 0.05$ (significant) or $\alpha = 0.01$ (highly significant).
+Metrics described in Appendix S1.
+Not all studies include super-endemism.
+
+- If either `pe_obs_p_upper` or `pe_alt_obs_p_upper` are significantly high (one-sided test), then look for paleo- or neo-endemism.
+  - If `rpe_obs_p_upper` is significantly high (two-sided test), then grid-cell is paleo-endemic.
+  - Else if `rpe_obs_p_lower` is significantly low (two-sided test), then grid-cell is neo-endemic.
+  - Else there is mixed age endemism, in which case:
+    - If both `pe_obs_p_upper` and `pe_alt_obs_p_upper` are highly significant (*p* < 0.01), then grid-cell is super-endemic.
+    - Else grid-cell is mixed-endemic.
+- Else if neither `pe_obs_p_upper` nor `pe_alt_obs_p_upper` are significantly high, grid-cell is non-endemic.

### 2022-10-13 ms/ms.Qmd
Update format for ecography

@@
 bibliography: references.yaml
-csl: methods-in-ecology-and-evolution.csl
+csl: ecography.csl
 highlight-style: github
@@
 
+{{< pagebreak >}}
+
+# canaper: Categorical analysis of neo- and paleo-endemism in R
+
 # Abstract
 
-1. Biodiversity has typically been quantified using species richness, but this ignores evolutionary history.
+Biodiversity has typically been quantified using species richness, but this ignores evolutionary history.
 Due to the increasing availability of robust phylogenies, methods have been developed that incorporate phylogenetic relationships into quantification of biodiversity.
@@
 
-2. Here, we present `canaper`, a new R package that provides functions to conduct CANAPE in R.
+Here, we present `canaper`, a new R package that provides functions to conduct CANAPE in R.
 `canaper` implements methods for efficient computation, including parallelization and encoding of community data as sparse matrices.
@@
 
-3. Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 10 min to reproduce a canonical study).
+Our case study shows that `canaper` produces equivalent results to Biodiverse and can complete computations on moderately sized datasets quickly (< 10 min to reproduce a canonical study).
 
-4.
 `canaper` allows researchers to conduct all analyses from data import and cleaning through CANAPE within R, thereby averting the need to manually import and export data and analysis results between programs.
@@
 
-Key-words: biodiversity, CANAPE, canaper, endemism, phylogeny, R, spatial phylogenetics
+Keywords: biodiversity, CANAPE, canaper, endemism, phylogeny, R, spatial phylogenetics
+
+{{< pagebreak >}}
 
@@
 
+<!-- FIXME: for double-blind review leave out of submission text, but add after accepted 
+To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 1.0.0':
+
+Nitta, J. H. et al. 2022. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD DETAILS HERE (ver 1.0.0) -->
+
+<!-- FIXME: for double-blind review leave out of submission text, but add after accepted 
 # Acknowledgements
 
-This research supported in part by Japan Society for the Promotion of Science (Kakenhi) Grant numbers 16H06279, 22H04925, and 22K15171.
 Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
 Luis Osorio and Klaus Schliep reviewed the code as part of submission to rOpenSci.
+-->
 
-# Conflict of interest
+# Funding
 
-All authors declare no conflict of interest.
+This research supported in part by Japan Society for the Promotion of Science (Kakenhi) Grant numbers 16H06279, 22H04925, and 22K15171.
 
@@
 
-# Data availability
+# Data availability statement
 
@@
 Usage of `canaper` is documented at <https://docs.ropensci.org/canaper/>.
+<!-- FIXME: for double-blind review leave out of submission text, but add after accepted 
 Code and data used to conduct analyses and generate this manuscript are available at <https://github.com/joelnitta/canaper_ms>, and can be run using the Docker image `joelnitta/canaper_ms` available at <https://hub.docker.com/r/joelnitta/canaper_ms>.
+-->
 The `acacia` dataset was originally published by @Mishler2014.

### 2022-10-14 ms/ms.Qmd
Update MS
fix some problems with refs, move Fig S1 code to SI

@@
 Joel H. Nitta  
-Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan  
 joelnitta@gmail.com
-
-Running headline: *canaper*
 
@@
 It is useful to plot the output to identify the minimum number of iterations needed.
-Here we do so with the `ggplot2` package.
-
-```{r}
-#| label: cpr-iter-sim-plot
-#| fig.height: 8
-#| eval: false
-
-library(ggplot2)
-
-ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
-  geom_line() +
-  labs(x = "Num. iterations", y = "% Similarity")
-```
-
-The plot shows that the community matrix becomes maximally randomized after about 40,000--50,000 iterations (@fig-iter-sim), and that the original matrix and randomized matrix are about 96.5% similar.
+Code to generate a plot with the `ggplot2` package and its output are shown in the online Supporting Information (Fig. S1).
+The plot shows that the community matrix becomes maximally randomized after about 40,000--50,000 iterations and that the original matrix and randomized matrix are about 96.5% similar.
 
@@
 
-a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
+a + b + plot_layout(ncol = 1) + 
+  plot_annotation(
+    tag_levels = "a",
+    tag_prefix = "(",
+    tag_suffix = ")")
 ```
@@
 Palettes for plotting CANAPE and *p*-rank results can be selected with the `cpr_make_pal()` function.
-The same plot visualized with CVD-safe colors is available in the online supporting material (Fig. S1).
+The same plot visualized with CVD-safe colors is available in Fig. S2.
 
@@
 <!-- FIXME: for double-blind review leave out of submission text, but add after accepted
+
 To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 1.0.0':
 
-Nitta, J. H. et al. 2022. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD DETAILS HERE (ver 1.0.0) -->
+Nitta, J. H. et al. 2022. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD DETAILS HERE (ver 1.0.0)
 
-<!-- FIXME: for double-blind review leave out of submission text, but add after accepted 
 # Acknowledgements
@@
 Luis Osorio and Klaus Schliep reviewed the code as part of submission to rOpenSci.
--->
 
@@
 
-Joel H.
-Nitta conceived the ideas and designed methodology; Joel H.
-Nitta and Shawn W.
-Laffan analyzed the data; Joel H.
-Nitta led the writing of the manuscript; Brent D.
-Mishler edited the manuscript; Wataru Iwasaki provided resources and supervised the study.
-All authors contributed critically to the drafts and gave final approval for publication.
+**Joel H.
+Nitta**: Conceptualization (Lead); Formal analysis (Lead); Funding acquisition (Supporting); Methodology (Lead); Software (Lead); Visualization (Lead); Writing – original draft (Lead); Writing – review & editing (Equal).
+**Shawn W. Laffan**: Data curation (Lead); Methodology (Supporting); Software (Supporting); Writing – review & editing (Equal).
+**Brent D. Mishler**: Writing – review & editing (Equal).
+**Wataru Iwasaki**: Funding acquisition (Lead); Project administration (Lead); Supervision (Lead); Writing – review & editing (Equal).
+
+-->
 
@@
 # References
-<!-- Insert <div id="refs"></div> to place the refs here instead of at the end -->
+
+<!-- FIXME: for some reason Lai2019 shows up as "in press" -->
 <div id="refs"></div>
@@
 For details on functions, see "Major Functions".
-
-**@fig-iter-sim.** Testing the number of iterations needed to randomize the Australian *Acacia* dataset.
-The *Acacia* dataset [@Mishler2014] was randomized using the `curveball` algorithm for 100,000 iterations and percentage similarity between the original matrix and the randomized matrix was calculated every 1,000 iterations using the `cpr_iter_sim()` function.
 
@@
 ```{r}
-#| label: fig-iter-sim
-#| fig.height: 4
-#| fig-cap: "Testing the number of iterations needed to randomize the Australian *Acacia* dataset."
-#| echo: false
-
-fig <- ggplot2::ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
-  geom_line() +
-  labs(x = "Num. iterations", y = "% Similarity")
-
-ggsave(plot = fig, file = here::here("ms/figure_2.pdf"),
-  width = 150, height = 150, units = "mm")
-
-fig
-```
-
-```{r}
 #| label: fig-acacia-cpr
@@
 
-fig <- a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
+fig <- a + b + plot_layout(ncol = 1) + plot_annotation(
+  tag_levels = "a",
+  tag_prefix = "(",
+  tag_suffix = ")")
 
-ggsave(plot = fig, file = here::here("ms/figure_3.pdf"),
+ggsave(plot = fig, file = here::here("ms/figure_2.pdf"),
   width = 100, height = 200, units = "mm")

### 2022-10-14 ms/si.Qmd
Update MS
fix some problems with refs, move Fig S1 code to SI

@@
 subtitle: "Supporting information"
-author:
-  - Joel H. Nitta
-  - Shawn W. Laffan
-  - Brent D. Mishler
-  - Wataru Iwasaki
+# FIXME: for double-blind review leave out of submission text, but add after accepted 
+# author:
+#   - Joel H. Nitta
+#   - Shawn W. Laffan
+#   - Brent D. Mishler
+#   - Wataru Iwasaki
 bibliography: references.yaml
@@
 ```
+
+```{r}
+#| label: iter-sim
+#| fig.height: 4
+#| echo: false
+
+set.seed(12345)
+
+data(acacia)
+
+# Conduct up to 100,000 iterations (swaps),
+# recording similarity every 1000 iterations
+iter_sim_res <- cpr_iter_sim(
+  comm = acacia$comm,
+  null_model = "curveball",
+  n_iterations = 100000,
+  thin = 1000
+)
+
+fig <- ggplot2::ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
+  geom_line() +
+  labs(x = "Num. iterations", y = "% Similarity")
+
+fig
+```
+
+**Fig S1.** Testing the number of iterations needed to randomize the Australian *Acacia* dataset.
+The *Acacia* dataset was randomized using the `curveball` algorithm for 100,000 iterations and percentage similarity between the original matrix and the randomized matrix was calculated every 1,000 iterations using the `cpr_iter_sim()` function.
+
+Code to generate Fig. S1 is as follows.
+
+```{r}
+#| label: cpr-iter-sim-show
+#| eval: false
+library(canaper)
+library(ggplot2)
+
+set.seed(12345)
+
+iter_sim_res <- cpr_iter_sim(
+  comm = acacia$comm,
+  null_model = "curveball",
+  n_iterations = 100000,
+  thin = 1000
+)
+
+ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
+  geom_line() +
+  labs(x = "Num. iterations", y = "% Similarity")
+```
+
+{{< pagebreak >}}
 
@@
 
-**Fig. S1**: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors.
+**Fig. S2**: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors (`cpr_endem_cols_4` palette in `canaper`).
 (a) Map of Australia showing grid-cells (communities) colored by endemism type.

### 2023-03-24 ms/ms.Qmd
Update ms according to reviewer comments

@@
 Quantifying biodiversity is a major goal of ecology.
-Quantification of biodiversity is often done on a purely taxonomic basis: counting the number of species in an area [e.g., @Diamond1975].
+The simplest and most commonly used biodiversity metric is species richness [e.g., @Diamond1975].
 However, as all taxa are related to some degree by descent from a common ancestor, a thorough understanding of biodiversity is only possible by considering their evolutionary relationships.
@@
 Community data may be either presence-absence data (0s or 1s) or abundance data (integers \>= 0).
-However, all calculations of PD and PE use only presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
+However, calculations of PD and PE for CANAPE use only presence-absence information (i.e., no abundance weighting is used), so identical results will be obtained whether the input data is abundance or abundance that has been converted to presence-absence.
 Community data is typically loaded using `read.csv()`, `readr::read_csv()`, or other functions that can import rectangular data.
@@
 Phylogenies should have no negative branch lengths, but are not required to be fully bifurcating.
+Either a phylogram or chronogram may be used.
 Phylogenies can be loaded with the `ape::read.tree()` function.
@@
 
-First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (`pd_obs`) and phylogenetic endemism (`pe_obs`) are calculated.
-Next, an alternative phylogeny (also referred to as the "comparison tree") is constructed that has non-zero branch lengths set to a constant value, then rescaled to a total length of 1.
+First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (`pd_obs`) and phylogenetic endemism (`pe_obs`) are calculated (Box 1).
+Next, an alternative phylogeny (also referred to as the "comparison tree") is constructed that has non-zero branch lengths set to a constant value (canaper arbitrarily uses 1), then rescaled to a total length of 1.
 PD and PE are then calculated on the alternative phylogeny (`pd_alt`, `pe_alt`).
-Relative PD and PE, the ratio of `pd_obs` to `pd_alt` and `pe_obs` to `pe_alt` respectively, are then calculated (RPD, RPE).
+Relative PD and PE, the ratio of `pd_obs` to `pd_alt` and `pe_obs` to `pe_alt` respectively, are then calculated (RPD, RPE; Box 1).
 However, the statistical significance of any of these metrics cannot be determined from observed values alone.
@@
 These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to one commonly used randomization algorithm in Biodiverse called `rand_structured` (see "Example: Australian *Acacia*").
+`curveball` has been shown to be more computationally efficient than `swap` as implemented in the `picante` package [@Strona2014], but we are unaware of any studies comparing the effects on PD or CANAPE between the two.
 
@@
 With a sufficiently high number of random communities, significance (e.g., at the $\alpha = 0.5$ level) is expected to converge, but there may be borderline cases that appear significant in some analyses and non-significant in others.
+(We have verified that calculation of deterministic metrics that do not rely on random values, e.g., raw PD, are identical between Biodiverse and `canaper`, and include a test for this in `canaper`).
 
@@
 Some of those that are more closely related to `canaper` include the following.
-`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000].
+`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000] and supports seven randomization algorithms, including two variations of the `swap` algorithm [@Gotelli2003; @Miklos2004].
 `phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing efficiency of PD, and is used by `canaper`.
 `vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
-`vegan` includes by far the greatest number of algorithms for generating random communities and is used by `canaper`.
+`vegan` includes by far the greatest number of algorithms for generating random communities, including `curveball` [@Strona2014] and both variations of `swap`, and is used by `canaper`.
 `adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.

### 2023-04-05 ms/ms.Qmd
Update MS

@@
 
-Joel H. Nitta^1,\*^, Shawn W. Laffan^2^, Brent D. Mishler^3^, and Wataru Iwasaki^1,4,5,6,7,8^
+Joel H. Nitta^1,9,\*^, Shawn W. Laffan^2^, Brent D. Mishler^3^, and Wataru Iwasaki^1,4,5,6,7,8^
 
@@
 ^8^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan
+
+^9^Current affiliation: College of Liberal Arts and Sciences, Graduate School of Global and Transdisciplinary Studies, Chiba University, Chiba, Japan
 
@@
 
-`canaper` allows researchers to conduct all analyses from data import and cleaning through CANAPE within R, thereby averting the need to manually import and export data and analysis results between programs.
+`canaper` allows researchers to conduct all analyses from data import and cleaning through CANAPE within R, thereby obviating the need to manually import and export data and analysis results between programs.
 We anticipate `canaper` will become a part of the toolkit for analyzing biodiversity in R.
@@
 This became possible with the development of phylogenetic measures of biodiversity, such as phylogenetic diversity [PD\; @Faith1992] and phylogenetic endemism [PE\; @Rosauer2009] (Box 1).
-Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies and large spatial datasets [e.g., @Mishler2020; @Nitta2022a; @Thornhill2017].
+Such analyses are becoming much more common due to the widespread availability of robust molecular phylogenies and large spatial datasets [e.g., @Thornhill2017; @Mishler2020; @Nitta2022a].
 
@@
 Until now, an R user who wanted to conduct CANAPE analysis as part of a broader R workflow needed to first clean raw data, export it to Biodiverse, conduct PD and PE analyses in Biodiverse, then import the results back into R for further analysis and visualization.
-Also, while Biodiverse calculated the metrics needed to categorize endemism type, it did not actually perform this categorization (although this feature will be included in a future version of Biodiverse).
-Furthermore, the Biodiverse GUI alone does not support parallel processing, which is needed for large datasets.
+Also, prior to version 4, Biodiverse calculated the metrics needed to categorize endemism type but did not actually perform this categorization.
+Furthermore, the Biodiverse GUI alone does not currently support parallel processing, which is important for large datasets.
 Parallel processing and automation of Biodiverse analyses is only currently possible using Perl scripts.
@@
 
-The `ape` R package is used to handle phylogenies, which are stored as lists of the class `phylo`.
-Phylogenies should have no negative branch lengths, but are not required to be fully bifurcating.
-Either a phylogram or chronogram may be used.
+The `ape` R package is used to handle phylogenies, which are stored as lists of the class `"phylo"`.
+Phylogenies are required to have branch lengths >= 0, but are not required to be fully bifurcating.
+Phylograms, chronograms, or cladograms may be used.
 Phylogenies can be loaded with the `ape::read.tree()` function.
@@
 
-First, the input phylogeny is scaled to a total length of 1 and observed phylogenetic diversity (`pd_obs`) and phylogenetic endemism (`pe_obs`) are calculated (Box 1).
-Next, an alternative phylogeny (also referred to as the "comparison tree") is constructed that has non-zero branch lengths set to a constant value (canaper arbitrarily uses 1), then rescaled to a total length of 1.
+First, the input phylogeny is scaled so the sum of all branch lengths is 1 and observed phylogenetic diversity (`pd_obs`) and phylogenetic endemism (`pe_obs`) are calculated (Box 1).
+Next, an alternative phylogeny (also referred to as the "comparison tree") is constructed that has non-zero branch lengths set to a constant value (`canaper` arbitrarily uses 1), then rescaled so the sum of all branch lengths is 1.
 PD and PE are then calculated on the alternative phylogeny (`pd_alt`, `pe_alt`).
@@
 
-PD (and by extension, PE, RPD, and RPE) will increase with taxon richness, since adding a new taxon adds new branches to the tree.
-To determine the statistical significance of these metrics, the observed value is compared to a distribution of values obtained from a set of random communities.
+PD (and by extension, PE) will increase with taxon richness, since adding a new taxon samples more branches of the tree.
+To determine the statistical significance of the calculated metrics, the observed value is compared to a distribution of values obtained from a set of random communities.
 The random communities are generated by a randomization algorithm that shuffles the original data.
@@
 These algorithms preserve the number of sites occupied by each species and the richness of each site, and both produce results comparable to one commonly used randomization algorithm in Biodiverse called `rand_structured` (see "Example: Australian *Acacia*").
-`curveball` has been shown to be more computationally efficient than `swap` as implemented in the `picante` package [@Strona2014], but we are unaware of any studies comparing the effects on PD or CANAPE between the two.
+`curveball` has been shown to be more computationally efficient than `swap`, but we are unaware of any studies comparing the effects on PD or CANAPE between the two.
 
@@
 The first argument, `comm`, is a community data frame (or matrix). 
-The second, `null_model`, is the name of one of the predefined randomization to use.
+The second, `null_model`, is the name of one of the predefined randomization algorithms to use.
 The remainder of the arguments are particular to specific types of randomization algorithm.
@@
 
-One feature to be aware of is that randomization algorithms in vegan are classified as either binary or quantitative.
+One feature to be aware of is that randomization algorithms in `vegan` are classified as either binary or quantitative.
 Binary algorithms are designed for binary (i.e., presence-absence) data, and quantitative algorithms are designed for quantitative (i.e., abundance) data.
@@
 The main arguments to this function are the input community and phylogeny, type of null model, and settings for the null model.
+Any species that do not appear in both the community data frame and the phylogeny will be excluded from analysis.
 For a full list of null models to choose from, run `?vegan::commsim()`.
@@
 To use parallel computing, no changes are needed for `cpr_rand_test()` etc.
-Rather, `future` is loaded with `library()`, then a parallel back-end is specified with `plan()`.
+Rather, `future` is loaded (`library("future")`), then a parallel back-end is specified with `plan()`.
 The user is advised to consult the `future` website (<https://future.futureverse.org/>) for more information on specifying a parallel backend.
@@
 With a sufficiently high number of random communities, significance (e.g., at the $\alpha = 0.5$ level) is expected to converge, but there may be borderline cases that appear significant in some analyses and non-significant in others.
-(We have verified that calculation of deterministic metrics that do not rely on random values, e.g., raw PD, are identical between Biodiverse and `canaper`, and include a test for this in `canaper`).
+Note that we have verified that calculation of deterministic metrics that do not rely on random values, e.g., raw PD, are identical between Biodiverse and `canaper`, and include a test for this in `canaper`).
 
@@
 `picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000] and supports seven randomization algorithms, including two variations of the `swap` algorithm [@Gotelli2003; @Miklos2004].
-`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing efficiency of PD, and is used by `canaper`.
+`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing efficiency of PD and PE, and is used by `canaper`.
 `vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
@@
 Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
+<!-- FIXME: Add new studies -->
 `canaper` has already been used at least three studies while in development [@Ellepola2022; @Lu2022; @Nitta2022a].
 We expect `canaper` will become a major tool in the toolkit of the emerging field of spatial phylogenetics alongside Biodiverse.
-
-<!-- FIXME: for double-blind review leave out of submission text, but add after accepted
-
-To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 1.0.0':
-
-Nitta, J. H. et al. 2022. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD DETAILS HERE (ver 1.0.0)
 
@@
 
--->
-
 # Data availability statement
@@
 Usage of `canaper` is documented at <https://docs.ropensci.org/canaper/>.
-<!-- FIXME: for double-blind review leave out of submission text, but add after accepted 
+
 Code and data used to conduct analyses and generate this manuscript are available at <https://github.com/joelnitta/canaper_ms>, and can be run using the Docker image `joelnitta/canaper_ms` available at <https://hub.docker.com/r/joelnitta/canaper_ms>.
--->
+
 The `acacia` dataset was originally published by @Mishler2014.
+
+To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 1.0.0':
+
+Nitta, J. H. et al. 2023. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD REFERENCE DETAILS HERE (version 1.0.0)
 

### 2023-04-07 _targets.R
Report progress of knitting

@@ -77,7 +77,8 @@ tar_plan(
       "ms/methods-in-ecology-and-evolution.csl",
       "ms/references.yaml",
       "ms/custom-reference.docx"
-    )
+    ),
+    quiet = FALSE
   ),
   # Render SI
   tar_quarto(
@@ -86,6 +87,7 @@ tar_plan(
     extra_files = c(
       "ms/methods-in-ecology-and-evolution.csl",
       "ms/references.yaml"
-    )
+    ),
+    quiet = FALSE
   )
 )

### 2023-04-07 ms/ms.Qmd
Update MS

@@
 To demonstrate usage with a real dataset, we reproduced the analysis of @Mishler2014, who conducted CANAPE on Australian *Acacia* using Biodiverse.
-All `canaper` analyses were run with `canaper` v`r packageVersion("canaper")` in R v`r R.Version()$major`.`r R.Version()$minor`.
+All `canaper` analyses were run with `canaper` v`r packageVersion("canaper")` in R v4.2.1.
 We used the `curveball` randomization algorithm with 50,000 iterations, since `cpr_iter_sim()` indicated this was sufficient to randomize the matrix (see "Major Functions").
@@
   mutate(across(c(long, lat), parse_number))
+min_long <- min(acacia_canape$long)
+min_lat <- min(acacia_canape$lat)
 
@@
   coord_fixed() +
+  # add a 1000 km scalebar (data units are in m)
+  geom_segment(
+    x = min_long,
+    y = min_lat,
+    xend = min_long + 1000 * 1000,
+    yend = min_lat
+  ) +
+  geom_text(
+    x = min_long,
+    y = min_lat,
+    label = "1,000 km",
+    vjust = -1, hjust = 0, size = 2
+  ) +
   theme(
@@
     legend.title = element_blank(),
-    panel.grid.minor = element_blank()
+    panel.grid.minor = element_blank(),
+    axis.title = element_blank(),
+    axis.text = element_blank(),
+    axis.ticks = element_blank()
   )
@@
 Furthermore, `canaper` features efficient computing routines and simple yet flexible implementation of parallel computing, thereby decreasing computation time.
-<!-- FIXME: Add new studies -->
-`canaper` has already been used at least three studies while in development [@Ellepola2022; @Lu2022; @Nitta2022a].
+`canaper` has already been used at least five studies while in development [@Ellepola2022; @Lu2022; @Nitta2022a; @Naranjo2023; @vanGalen2023].
 We expect `canaper` will become a major tool in the toolkit of the emerging field of spatial phylogenetics alongside Biodiverse.
@@
 
-To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 1.0.0':
+To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 0':
 
-Nitta, J. H. et al. 2023. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD REFERENCE DETAILS HERE (version 1.0.0)
+<!-- FIXME: update journal details when available -->
+Nitta, J. H. et al. 2023. canaper: Categorical analysis of neo- and paleo-endemism in R. Ecography XX: XXX--XXX (version 0)
 
@@
 (a) Map of Australia showing grid-cells (communities) colored by endemism type.
-Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system [@Butler2007]; units in m.
+Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system [@Butler2007].
 (b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.
@@
 
+min_long <- min(acacia_canape$long)
+min_lat <- min(acacia_canape$lat)
+
 a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
   geom_tile() +
-  # use same colors as Misher et al. 2014
-  # (for CVD-safe colors could use cpr_endem_cols or cpr_endem_cols_2)
+  # Use same colors as Misher et al. 2014
+  # (for CVD-safe colors see cpr_endem_cols)
   scale_fill_manual(values = mishler_endem_cols) +
@@
   coord_fixed() +
+  # add a 1000 km scalebar
+  geom_segment(
+    x = min_long,
+    y = min_lat,
+    xend = min_long + 1000 * 1000,
+    yend = min_lat
+  ) +
+  geom_text(
+    x = min_long,
+    y = min_lat,
+    label = "1,000 km",
+    vjust = -1, hjust = 0, size = 2
+  ) +
   theme(

### 2023-04-07 ms/si.Qmd
Update MS

@@
 
+min_long <- min(acacia_canape$long)
+min_lat <- min(acacia_canape$lat)
+
 a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
@@
     panel.grid.major = element_line(color = "grey60"),
-    panel.grid.minor = element_blank()
+    panel.grid.minor = element_blank(),
+    axis.title = element_blank(),
+    axis.text = element_blank(),
+    axis.ticks = element_blank()
   )
@@
 (a) Map of Australia showing grid-cells (communities) colored by endemism type.
-Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system; units in m.
+Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system.
 (b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.

### 2023-05-14 ms/ms.Qmd
Update ref format don't shorten et al list in refs

@@
 bibliography: references.yaml
-csl: ecography.csl
+csl: oikos.csl
 highlight-style: github

### 2023-05-14 ms/si.Qmd
Update ref format don't shorten et al list in refs

@@
 bibliography: references.yaml
-csl: methods-in-ecology-and-evolution.csl
+csl: oikos.csl
 highlight-style: github

### 2023-05-14 _targets.R

@@ -74,7 +74,7 @@ tar_plan(
     ms,
     "ms/ms.Qmd",
     extra_files = c(
-      "ms/methods-in-ecology-and-evolution.csl",
+      "ms/oikos.csl",
       "ms/references.yaml",
       "ms/custom-reference.docx"
     ),
@@ -85,7 +85,7 @@ tar_plan(
     si,
     "ms/si.Qmd",
     extra_files = c(
-      "ms/methods-in-ecology-and-evolution.csl",
+      "ms/oikos.csl",
       "ms/references.yaml"
     ),
     quiet = FALSE

### 2024-05-29 ms/ms.Qmd
Fix affiliations

@@
 
-^2^School of Biological, Earth and Environmental Sciences, The University of New South Wales, Kensington, Australia
+^2^Earth and Sustainability Science Research Centre, School of Biological, Earth and Environmental Sciences, The University of New South Wales, Sydney, NSW, Australia
 
-^3^Department of Integrative Biology, University of California Berkeley, Berkeley, USA
+^3^Jepson and University Herbaria, and Department of Integrative Biology, University of California Berkeley, Berkeley, USA
 
@@
 In this case, the data must include both row names and column names.
+Metrics of phylogenetic diversity including CANAPE can be measured at any taxonomic level, but for simplicity we use ’species’ here.
 Community data may also be input as a tibble, in which case site names must be indicated in a dedicated column (default column name `"site"`), rather than row names since tibbles lack row names.
@@
 Since the randomization algorithm influences the range of reference (expected) values, the choice of randomization algorithm is likely to have a large effect on the results.
-As there is no single "correct" algorithm, we have opted to provide the user with a wide range of options by implementing randomization algorithms included in the `vegan` package.
+As there is no single "correct" algorithm, we have opted to provide the user with a wide range of options by supporting randomization algorithms included in the `vegan` package.
 `vegan` includes \>30 randomization algorithms, but not all are appropriate for CANAPE.
@@
 A number slightly larger than the smallest number of iterations needed to reach the approximate minimum value can then be used for randomizing the community.
-The following code demonstrates usage of `cpr_rand_test()` with the *Acacia* dataset [@Mishler2014] that comes with `canaper`.
+The following code demonstrates usage of `cpr_iter_sim()` with the *Acacia* dataset [@Mishler2014] that comes with `canaper`.
 The *Acacia* dataset is relatively large (@tbl-datasets) and highly skewed (`r acacia_non_empty` zeros).
@@
 `acacia` includes data of Australian *Acacia* analyzed by @Mishler2014.
-`phylocom` and `biod_example` are fictitious datasets compiled for testing software, and were originally published in Phylocom [@Webb2008] and Biodiverse [@Laffan2010], respectively.
+`phylocom` and `biod_example` are synthetic datasets compiled for testing software, and were originally published in Phylocom [@Webb2008] and Biodiverse [@Laffan2010], respectively.
 
@@
 With a sufficiently high number of random communities, significance (e.g., at the $\alpha = 0.5$ level) is expected to converge, but there may be borderline cases that appear significant in some analyses and non-significant in others.
-Note that we have verified that calculation of deterministic metrics that do not rely on random values, e.g., raw PD, are identical between Biodiverse and `canaper`, and include a test for this in `canaper`).
+Note that we have verified that calculation of deterministic metrics that do not rely on random values, e.g., raw PD, are identical between Biodiverse and `canaper`, and include a test for this in `canaper`.
 
@@
 Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
-Luis Osorio and Klaus Schliep reviewed the code as part of submission to rOpenSci.
+Luis Osorio and Klaus Schliep reviewed the code as part of submission to rOpenSci. Three anonymous reviewers provided comments that improved the manuscript.
 
@@
 Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system [@Butler2007].
-(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.
+(b) Scatterplot comparing phylogenetic endemism (PE) of each community as measured on the original tree vs.
 a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type.
